\documentclass[handout]{ximera}

\input{../preamble}

\pgfplotsset{compat=1.13}

\outcome{Compute complex integrals around closed curves}

\title{5.2 Cauchy's Theorem}

\begin{document}

\begin{abstract}
We compute integrals of complex functions around closed curves.
\end{abstract}

\maketitle

A contour is called closed if its initial and terminal points coincide.  Further it is called simple if those are the only points which coincide. 
\begin{theorem}[Cauchy's Theorem]
If $f$ is analytic along a simple closed contour $C$ and also analytic inside $C$, then
\[
\int_C f(z) \, dz = 0
\]
\end{theorem}

\begin{example}
Let $C$ be a simple closed contour that does not pass through $z_0$ or contain $z_0$ in its interior. Compute $\dis \int_C \frac{1}{z-z_0} \, dz$.\\
The function $\dis f(z) = \frac{1}{z-z_0}$ is analytic everywhere except at $z_0$.  From the description of the contour $C$,
$f(z)$ is analytic along $C$ and in its interior.  Hence, by Cauchy's Theorem,
\[
\int_C \frac{1}{z-z_0} \, dz = 0
\]
\end{example}
\begin{remark}
If the contour $C$ in the previous example passed through the point $z_0$ or contained $z_0$ in its interior, Cauchy's Theorem would not apply
and the integral could be non-zero.
\end{remark}

\begin{problem}(problem 1a)
Let $C = C(z_0, r)$ traversed clockwise.\\
\[
\int_C e^z \, dz = \answer{0}
\]
\end{problem}

\begin{problem}(problem 1b)
Let $C = C(z_0, r)$ traversed clockwise.\\
\[
\int_C \sin(z) \, dz = \answer{0}
\]
\end{problem}

\begin{problem}(problem 1c)
Let $C = C(1+i, 1)$ traversed counter-clockwise.\\
\[
\int_C \Log(z) \, dz = \answer{0}
\]
\end{problem}

\begin{problem}(problem 1d)
Let $C = C(1+i, 1)$ traversed counter-clockwise.\\
\[
\int_C \frac{1}{z} \, dz = \answer{0}
\]
\end{problem}



A \textbf{positively oriented} simple closed contour is one in which the interior of the contour is to the left as the contour is traced out.
In the case of a circle, positive orientation is counter-clockwise. 

\begin{theorem}[Deformation of Contour]
Suppose $C_1$ and $C_2$ are positively oriented, simple closed contours such that $C_1$ lies entirely inside $C_2$. If $f$ is analytic 
along $C_1$ and $C_2$ and also analytic in the region between them, then
\[
\int_{C_1} f(z) \, dz = \int_{C_2} f(z) \, dz
\]
\end{theorem}

\begin{remark} 
The Deformation of Contour theorem is typically used to replace arbitrary positively oriented, simple closed curves with circles traversed counter-clockwise.
\end{remark}


\begin{example}[example 2]
Let $C$ be a positively oriented, simple closed contour containing the origin in its interior. Compute $\dis \int_C \frac{1}{z} \, dz$.\\
Since the origin is interior to $C$, there exists a radius $r$ such that the circle $C(0, r)$ lies in the interior of $C$.
By the Deformation of Contour theorem,
\[
\int_C \frac{1}{z} \, dz = \int_{C(0,r)} \frac{1}{z} \, dz
\]
where $C(0,r)$ is traversed in the positive direction, i.e., counter-clockwise. Using the parametrization
$\gamma(t) = re^{it}, \, 0\leq t \leq 2\pi$ for $C(0, r)$, we have
\[
\int_{C(0,r)} \frac{1}{z} \, dz = \int_0^{2\pi} \frac{\gamma'(t)}{\gamma(t)} \, dt = \int_0^{2\pi} i \, dt = 2\pi i
\]
Hence, 
\[
\int_C \frac{1}{z} \, dz = 2\pi i
\]
as well.
\end{example}

\begin{problem}(problem 2a)
Let $C$ be a positively oriented simple closed curve containing the point $z_0$ in its interior.\\
\[
\int_C \frac{1}{z-z_0} \, dz = \answer{2\pi i}
\]
\end{problem}

\begin{problem}(problem 2b)
Let $C$ be a positively oriented simple closed curve containing the point $z_0$ in its interior.\\
\[
\int_C \frac{1}{(z-z_0)^2} \, dz = \answer{0}
\]
\end{problem}


\begin{problem}(problem 2c)
Let $C$ be a positively oriented simple closed curve containing the point $z_0$ in its interior and let $n > 2$.\\
\[
\int_C \frac{1}{(z-z_0)^n} \, dz = \answer{0}
\]
\end{problem}


Here is a video solution to problem 2c:\\
\begin{foldable}
\youtube{RrtfW4YuM5M}
\end{foldable}

\begin{theorem}[Extended Deformation of Contour]
Suppose $C_1, C_2, ..., C_n$ are non-intersecting positively oriented, simple closed contours each lying inside of another positively oriented, simple closed contour $C$.  
If $f$ is analytic along $C, C_1, C_2, ..., C_n$ and also analytic in the region between them (i.e. interior to $C$ but exterior to each of $C_1, C_2, ..., C_n$), then
\[
\int_C f(z) \, dz = \sum_{k=1}^n \int_{C_k} f(z) \, dz
\]
\end{theorem}



\begin{theorem}[Cauchy's Integral Formula]
Suppose $f$ is analytic along and inside a positively oriented, simple closed contour $C$ containing the point $z_0$ in its interior.
Then
\[
f^{(n)}(z_0) = \frac{n!}{2\pi i} \int_C \frac{f(z)}{(z-z_0)^{n+1}} \, dz
\]
\end{theorem}

\begin{remark}
The case $n=0$ corresponds to 
\[
f(z_0) = \frac{1}{2\pi i} \int_C \frac{f(z)}{z-z_0} \, dz
\]
\end{remark}


\begin{example}[example 3a]
Let $C = C(0, 1)$ traversed counter-clockwise. Compute the contour integral: $\dis \int_C \frac{\sin z}{z^2} \, dz$\\
We use Cauchy's Integral Formula.  To do so, we first need to identify $z_0, n$ and $f(z)$ and verify that $f(z)$ is analytic in an on $C$.
We have $z_0 = 0, n=1, $ and $f(z) = \sin z$ which is entire. Thus, we can apply the formula and we obtain
\[
\int_C \frac{\sin z}{z^2} \, dz = \frac{2\pi i}{1!} f'(0) = 2\pi i \cos 0  = 2 \pi i
\]
\end{example}

\begin{example}[example 3b]
Let $C = C(2, 1)$ traversed counter-clockwise. Compute the contour integral: $\dis \int_C \frac{\sin z}{z(z-2)} \, dz$\\
We use Cauchy's Integral Formula.  To do so, we first need to identify $z_0, n$ and $f(z)$ and verify that $f(z)$ is analytic in an on $C$.
We have $z_0 = 2, n=0, $ and $f(z) = \frac{\sin z}{z}$ which is analytic in and on $C(2,1)$. Thus, we can apply the formula and we obtain
\[
\int_C \frac{\sin z}{z(z-2)} \, dz = \frac{2\pi i}{0!} f(2) = 2\pi i \frac{\sin 2}{2}  =  \pi i \sin 2
\]
\end{example}


\begin{problem}(problem 3a)
Let $C = C(0, 1)$ traversed counter-clockwise. Compute the contour integral: $\dis \int_C \frac{e^z}{z^2} \, dz$\\
\[
z_0 = \answer{0} \quad n = \answer{1} \quad f(z) = \answer{e^z}
\]
\[
\int_C \frac{e^z}{z^2} \, dz = \answer{2\pi i}
\]
\end{problem}
\begin{problem}(problem 3b)
Let $C = C(0, 1)$ traversed counter-clockwise. Compute the contour integral: $\dis \int_C \frac{e^z}{z^4} \, dz$\\
\[
z_0 = \answer{0} \quad n = \answer{3} \quad f(z) = \answer{e^z}
\]
\[
\int_C \frac{e^z}{z^4}\, dz = \answer{i pi /3}
\]
\end{problem}
\begin{problem}(problem 3c)
Let $C = C(i, 1)$ traversed clockwise. Compute the contour integral: $\dis \int_C \frac{\cos z}{z-i} \, dz$\\
\[
z_0 = \answer{i} \quad n = \answer{0} \quad f(z) = \answer{\cos z}
\]
\[
\int_C \frac{\cos z}{z-i}\, dz = \answer{-2\pi i \cosh 1}
\]
\end{problem}
\begin{problem}(problem 3d)
Let $C = C(4, 2)$ traversed counter-clockwise. Compute the contour integral: $\dis \int_C \frac{\sin z}{(z-1)(z-3)} \, dz$\\
\[
z_0 = \answer{3} \quad n = \answer{0} \quad f(z) = \answer{\frac{\sin z}{z-1}}
\]
\[
\int_C \frac{\sin z}{(z-1)(z-3)} \, dz = \answer{\pi i \sin 3}
\]
\end{problem}

Here is a video solution to problem 3b:\\
\begin{foldable}
\youtube{_FgmP-khmiQ}
\end{foldable}


\begin{example}[example 4]
Let $C = C(0, 2)$ traversed counter-clockwise. Compute the contour integral: $\dis \int_C \frac{e^z}{z^2 +1} \, dz$\\
The integrand has singularities at $\pm i$, so we use the Extended Deformation of Contour Theorem before we use Cauchy's Integral Formula.
By the Extended Deformation of Contour Theorem we can write
\[
\int_C \frac{e^z}{z^2 +1} \, dz = \int_{C_1} \frac{e^z}{z^2 +1} \, dz +  \int_{C_2} \frac{e^z}{z^2 +1} \, dz 
\]
where $C_1 = C(i, 1/2)$ traversed counter-clockwise and $C_2 = C(-i, 1/2)$ traversed counter-clockwise.
Now by Cauchy's Integral Formula with $n = 0$, we have
\[
\int_{C_1} \frac{e^z}{z^2 +1} \, dz = \int_{C_1} \frac{e^z}{(z+i)(z-i)} \, dz = 2\pi i f(i)
\]
where $f(z) = \frac{e^z}{z+i}$.
Hence
\[
\int_{C_1} \frac{e^z}{z^2 +1} \, dz = 2\pi i \cdot \frac{e^i}{2i} = \pi e^i
\]
Similarly,
\[
\int_{C_2} \frac{e^z}{z^2 +1} \, dz = \int_{C_2} \frac{e^z}{(z+i)(z-i)} \, dz = 2\pi i f(-i)
\]
where $f(z) = \frac{e^z}{z-i}$.
Hence
\[
\int_{C_1} \frac{e^z}{z^2 +1} \, dz = 2\pi i \cdot \frac{e^{-i}}{-2i} = -\pi e^{-i}
\]
Combining these we get
\[
\int_C \frac{e^z}{z^2 +1} \, dz = \pi e^i - \pi e^{-i} = 2\pi \sin(1)
\]
\end{example}

\begin{problem}(problem 4)
Let $C = C(0, 2)$ traversed counter-clockwise. Compute the contour integral:\\
\[
\int_C \frac{e^z}{z^2 -1} \, dz  \answer{\pi i (e-1/e)}
\]
\end{problem}

Here is a video solution to problem 4:\\
\begin{foldable}
\youtube{MJ50LDr2bUg}
\end{foldable}

\end{document}






