\documentclass[handout]{ximera}
% \usepackage{tcolorbox}
\input{../preamble}

\outcome{Compute the derivative of a composition}

\title{2.15 Rates of Change}



\begin{document}

\begin{abstract}
In this section we interpret the derivative as an instantaneous rate of change.
\end{abstract}

\maketitle

\section{Rates of Change}

We begin by comparing the notion instantaneous rate of change to average rate of change. 
For two points $(x_1, y_1)$ and $(x_2, y_2)$ with $x_1 < x_2$ on the graph of a function
$y = f(x)$, the slope of the secant line connecting them is 
\[m = \frac{\text{rise}}{\text{run}} = \frac{\Delta y}{\Delta x} = \frac{y_2 - y_1}{x_2 - x_1}.\]
This slope can be interpreted as the average rate of change of $y$ with respect to $x$ over the interval $[x_1, x_2]$.

The instantaneous rate of change of $y$ with respect to $x$ at $x = x_1$ is denoted by
\[\frac{dy}{dx}\Big{|}_{x=x_1},\]
and it is obtained by letting $x_2$ approach $x_1$ in the formula for $\frac{\Delta y}{\Delta x}.$
Graphically, 
\[\frac{dy}{dx}\Big{|}_{x=x_1}\]
is the slope of the tangent line to the graph of $y = f(x)$ at $x=x_1$.
Generally speaking, $\frac{\Delta y}{\Delta x}$
represents the average rate of change of $y$ relative to $x$ whereas
$\frac{dy}{dx}$
represents the instantaneous rate of change of $y$ relative to $x$.

%More generally, $\frac{dy}{dx}$ is a formula for the instantaneous rate of change of $y$ with respect to $x$.

\begin{center}
\bf{Examples of Rates of Change}
\end{center}
 
\begin{example}[example 1]

If $y = f(t)$ is the distance traveled by an object as a function of time, then 
\[\frac{\Delta y}{\Delta t}\]
is the average speed of the object and
\[\frac{dy}{dt}\]
is the instantaneous speed of the object.
\end{example}

\begin{example} If $V = f(t)$ is the volume of water in a tank as a function of time, then 
\[\frac{\Delta V}{\Delta t}\]
is the average rate that water is entering (or leaving, if this is negative) the tank and
\[\frac{dV}{dt}\]
is the instantaneous rate that water is entering (or leaving) the tank.
 \end{example}

\begin{example}[example 2]
 If $C = f(x)$ is the cost of producing $x$ units then $\frac{dC}{dx}$ represents the instantaneous rate of change 
of cost with respect to the number of units produced.  This is called the {\it marginal cost} of production. 
\end{example}


\begin{example}[example 3]
 If $P(t)$ represents population at time $t$, then 
\[\frac{\Delta P}{\Delta t}\]
is the average growth rate of the population and
\[\frac{dP}{dt}\]
is the instantaneous growth rate of the population. If the population of a certain bacteria is 50 (thousand) and it doubles every hour, 
then what is the population after 3 hours?  How fast is the population growing at $t = 3$ hours? First, from the description of the population, 
we have $P(t) = 50 \cdot 2^t$.  Hence the population after 3 hours is given by $P(3) = 50\cdot 2^3 = 400$ (thousand) bacteria.
The growth rate at time $t=3$ is given by
\[\frac{dP}{dt}\Big{|}_{t=3}.\]
Computing the derivative, we have 
\[\frac{dP}{dt}= 50 \cdot 2^t \cdot \ln(2),\]
and hence,
\[\frac{dP}{dt}\Big{|}_{t=3} = 50 \cdot 2^3 \cdot \ln(2) = 400\ln(2) \approx 277.26 \mbox{(thousand) bacteria per hour}.\]
\end{example}

\begin{example}[example 4]
 If $Q(t)$ represents the quantity of a (medical) drug in the bloodstream at time $t$, then 
\[\frac{\Delta Q}{\Delta t}\]
is the average rate of change of the quantity in the bloodstream and
\[\frac{dQ}{dt}\]
is the instantaneous rate of change of the quantity in the bloodstream.


\end{example}

\end{document}
