\documentclass[handout]{ximera}

\input{../preamble}

\pgfplotsset{compat=1.13}

\outcome{Define and enumerate combinations}

\title{1.5 Combinations}

\begin{document}

\begin{abstract}
We define and enumerate combinations.
\end{abstract}

\maketitle

\section{Combinations}



\begin{definition}[Combination]
A combination is a selection of objects in which the order of selection is unimportant.
\end{definition}

Combinations are counted by counting permutations and then dividing appropriately.


\begin{example}[example 1]
In how many ways can a 6 person committee be formed from a department of 30 faculty members?\\
The order in which the members are selected is not important, so the answer is the number of combinations of 30 objects taken 6 at a time, denoted $C(30,6)$.  To count this, we note that each 6 person committee can have its members permuted in $6!$ ways. Thus $6! \cdot C(30,6) = P(30,6)$.  Hence,
\[
C(30,6) = \binom{30}{6} = \frac{P(30,6)}{6!} = \frac{30!}{6!24!} = 593,775
\]
\end{example}

\begin{problem}(problem 1)
In how many ways can a 5 person committee be formed from a group of 25 employees? $\; \answer{53130}$\\
\end{problem}

\begin{example}[example 2]
How many subsets of size 8 can be formed from a set of size 12?\\
Since the order of selection of the 8 elements in the subset is not important, the answer is the number of combinations of 12 objects taken 8 at a time, denoted $C(12,8)$.  To count this, we note that each 8 element subset can have its members permuted in $8!$ ways. Thus $8! \cdot C(12,8) = P(12,8)$.
\[
C(12,8) = \binom{12}{8} = \frac{P(12,8)}{8!} = \frac{12!}{8!4!} =  495
\]
\end{example}
\begin{proposition}
The number of combinations of $n$ objects taken $k$ at a time is
\[
C(n,k) = \binom{n}{k} = \frac{P(n,k)}{k!} = \frac{n!}{k!(n-k)!}
\]

\end{proposition}
\begin{proof}
Each combination of the $k$ objects selected from a set of $n$ distinct objects 
corresponds to $k!$ permutations of those $k$ objects.  
This is because there are $k!$ ways to permute the $k$ selected objects.
Thus the number of permutations, $P(n,k)$ is $k!$ times greater than the 
number of combinations, $C(n,k)$. Since we know the number of permutations is 
\[
P(n,k) = \frac{n!}{(n-k)!}
\]
we have
\[
P(n,k) = k!\cdot C(n,k)
\]
Dividing both sides by $k!$ gives the desired result, namely
\[
C(n,k) = \frac{n!}{k!(n-k)!}
\]


\end{proof}


\begin{example}[example 3]
Compute $C(7, 3)$.\\
Using the formula from the proposition above, with $n =7$ and $k = 3$, we have
\[
C(7,3) = \frac{7!}{3!4!} = \frac{7 \cdot 6\cdot 5}{3\cdot 2\cdot 1} = 35
\]
\end{example}


\begin{problem}(problem 3)
Compute each of the following:\\
$C(5,2) = \; \answer{10}$\\
$C(8,3) = \; \answer{56}$\\
$C(10,5) = \; \answer{252}$\\
$C(n,n) = \; \answer{1}$
\end{problem}



\begin{example}[example 4]
An NBA basketball team consists of 15 players.  In how many ways can a coach choose the 
starting 5 (without regard to position, ability, etc)?\\
The order in which the 5 players are selected is not important, so the answer is $C(15, 5)$.
\end{example}

\begin{problem}(problem 4)
An NHL hockey team consists of 23 players.  A line-up consists of 6 players.  
How many line-ups are possible (without regard to position)? $\; \answer{100947}$\\

\end{problem}


\begin{example}[example 5]
How many subsets of size $k$ can be made from a set containing $n$ elements ($k\leq n$)?\\
\end{example}


\begin{problem}(problem 5)
How many subsets of size $n-k$ can be made from a set containing $n$ elements ($k\leq n$)?\\
How does this answer differ from the answer to the example above?\\
\end{problem}


\begin{example}[example 6]
A box of crayons contains 64 different colors.  In how many ways can Jack select
 6 different colors to make his drawing?\\
 Since the order of selection of the 6 colors is not important, there are 
 \[
 C(64, 6) = \binom{64}{6} = \frac{64!}{6!58!}=74974368
 \]
 ways to select the 6 colors from a box containing 64 different colors.
 
\end{example}


\begin{problem}(problem 6)
A box of crayons contains 64 different colors.  In how many ways can Jill select
 10 different colors to make her drawing? $\; \answer{151473214816}$\\
\end{problem}



\begin{example}[example 7] 
An ordinary deck of playing cards contains 52 cards broken into two colors, red and black (26 of each).  
Each color is further subdivided into two suits (of 13 cards each)- red are evenly split between 
hearts and diamonds, black between spades and clubs. 
Finally, each suit is decomposed into 13 types labeled 2 through 10, then Jack, Queen, King and Ace.
In the popular card game poker, each player is dealt 5 cards.  How many different 5 card hands are possible?\\
The answer is simply $C(52, 5)$ since the order in which the cards are received is 
irrelevant (and quickly becomes unknown when the holder organizes the cards in their hand).
\end{example}


\begin{problem}(problem 7)
There are variants of poker in which a player is dealt 7 cards. How many $7$-card poker hands are possible? $\; \answer{\frac{(52)!}{7!(45)!}}$
\end{problem}

\begin{example}[example 8] 
In poker, one type of hand is called ``three of a kind". This occurs when 3 of your 5 cards are of 
the same type and the remaining two cards are of two other types.  For example, a hand consisting of 3 Jacks, an 
Eight and a Four (regardless of suit) is a three of a kind (3 Jacks). How many three of a kinds are possible?\\
We can break this down into a two step process.  First we choose the 3 cards to make the ``three of a kind", 
then choose the remaining two cards. To choose the 3 matching cards, there are 13 choices for the type and 
then there are $C(4, 3) =4$ ways to choose the suits.  To choose the remaining two cards, 
we first choose two different types in $C(12, 2)$ ways (one of the 13 types is accounted for, 
so there are only 12 left), and then choose the suit of each card in $4^2$ ways.
Putting all of this together, the number of "three of a kind's" is
\[
13 \cdot 4 \cdot C(12, 2) \cdot 4^2 = 54, 912
\]

\end{example}

\begin{problem}(problem 8)
How many of each of the following poker hands are possible?
a) Four of a kind: $\; \answer{624}$\\
b) Flush (all 5 cards of the same suit): $\; \answer{5148}$\\
c) Full house (three of one kind and two of another kind): $\; \answer{3744}$\\
d) Two pair: $\; \answer{123552}$
\end{problem}


\end{document}









