\documentclass[handout]{ximera}

\input{../preamble}

\pgfplotsset{compat=1.13}

\outcome{Define and enumerate circular permutations}

\title{1.6 Circular Permutations}

\begin{document}

\begin{abstract}
We define and enumerate circular permutations.
\end{abstract}

\maketitle

\section{Circular Permutations}



\begin{definition}[Circular Permutation]
A circular permutation of selection of objects placed in a circular 
arrangement in a particular order.
\end{definition}


In a circular permutation, all positions on the circle are considered equivalent. 
Thus, the position of an object
is solely determined by its position relative to the other objects. 
By contrast, the objects in an ordinary permutations have 
absolute positions- first, second, third etc.  An ordinary permutation 
can be thought of as a linear permutation.\\

\begin{proposition}
The number of circular permutations of $k$ objects taken from a 
group of $n$ objects ($n \geq k$) is given by
\[
C(n,k)\cdot (k-1)! = \frac{P(n,k)}{k}
\]
\end{proposition}

\begin{proof}
First, we select the $k$ objects to be placed in the circular permutation.  
This can be done $C(n,k)$ ways.\\
Second, we arrange the $k$ objects in a circle and use the FPC. 
When the first object is placed in the circle, all of the positions are 
equivalent, so there is only 1 choice. Once the first object is placed, 
the remaining positions in the circle become distinct.  
Hence, there are $k-1$ choices for the position of the second object in the 
circle, $k-2$ choices for the third object, and so on. 
Thus, the number of ways to arrange the $k$ objects in a circle  is $(k-1)!$.
Finally, we multiply the number of ways to select the $k$ objects by the 
number of ways to arrange them in a circle to obtain
\[
C(n,k)\cdot (k-1)!
\]
To see that this is equal to $\frac{P(n,k)}{k}$, we note that 
\[
\frac{(k-1)!}{k!} = \frac{1}{k}
\]
\end{proof}


\begin{corollary}
The number of circular permutations of $n$ objects (taken $n$ at a time) is $(n-1)!$
\end{corollary}
\begin{proof}
Applying the proposition with $k = n$ yields that the number of such permutations is
\[
C(n,n)\cdot (n-1)! = (n-1)!
\]
\end{proof}

\begin{example}[example 1]
In how many ways can 10 first graders march in a circle?\\
Here, we are counting the number of circular permutations of $10$ 
objects (taken $10$ at a time).
According to the corollary, the number of ways for the 10 children to 
march in a circle is $9!$.
\end{example}


\begin{problem}(problem 1)
In how many ways can the 20 members of the high-school band 
march in a circle during the halftime show? $\; \answer{19!}$
\end{problem}


\begin{example}[example 2]
How many ways can King Arthur, Guinevere, Merlin, Lancelot and Galahad be 
seated at the round table if Merlin and Galahad cannot be seated next to each other?\\
According to the corollary, the total number of ways for 
them to be seated around the table with no restrictions is $4! =24$\\
Next, we count the number of seating arrangements that have Merlin and 
Galahad next to each other.  We seat them first, in 2 possible 
ways- with Galahad on Merlin's left and with Galahad on Merlin's right. 
Then the other 3 members can be seated in 3! = 6 ways.  
Thus there are $2 \cdot 6 = 12$ ways for the group to be seated in 
violation of the restriction.\\
Finally, by subtracting (using the Rule for Complements), we conclude 
that the number of ways to seat the group so that Merlin and Galahad are 
separated is $24-12 = 12$.
\end{example}


\begin{problem}(problem 2)
In how many ways can the President, Vice-President, Secretary of State, 
Secretary of the Treasury, Secretary of the Interior, Secretary of Labor 
and a secretary be seated at a round table if the President and Vice-President 
cannot be seated next to each other? $\; \answer{480}$.

\end{problem}


\begin{example}[example 3]
In how many ways can 8 children be seated around a table at a birthday 
party if they sit boy-girl-boy-girl?\\
The first child, say a girl is seated in 1 way, since all seats at a round 
table are considered identical. Once she is seated, there are 3 seats left 
for the other 3 girls and therefore they can be seated in $3!$ ways.  
The 4 boys can then be seated in the remaining 4 seats in $4!$ ways.\\
The total number of ways to seat the eight children is 
\[
1\cdot 3! \cdot 4! = 6 \cdot 24 = 144
\]

\end{example}


\begin{problem}(problem 3)
In how many ways can 8 teenagers sit at a round table if the 4 boys sit next to each other?
$\; \answer{576}$
\end{problem}


\begin{example}[example 4]
Twenty children run to get in line to get on the carousel when the local amusement 
park opens its doors at 9am. If the ride has 10 seats, alternating between 
horses and tigers, how many ways are there for the first group of 10 
children to be seated on the ride?\\
First, we have to choose the 10 children to be seated.  This can be done in $C(20,10)$ ways.
Next, we can seat the first child in 2 ways (horse or tiger) and then the other 9 children 
can be seated in $9!$ ways relative to the first child.\\
Hence, the total number of configurations of 10 of the 20 children for the 
first carousel ride of the day is 
\[
C(20,10) \cdot 2 \cdot 9! = \frac{20! \cdot 2 \cdot 9!}{10!\cdot 10!} = 134088514560
\]
\end{example}


\begin{problem}(problem 4)
Fifteen children run to get in line to get on the carousel when the local amusement 
park opens its doors at 10am. The ride has 12 seats arranged in a circle. 
The seats alternate horse, horse, tiger, tiger, etc. 
How many ways are there for the first group of 12 
children to be seated on the ride? $\; \answer{72648576000}$

\end{problem}


\begin{example}[example 5]
Sam is making a bracelets consisting of 8 beads, each of a different color. 
How many different looking bracelets can Sam create?\\
At first, this may seem like a straight forward circular permutation, 
but unlike the other scenarios we have seen, the round object in this 
case can be flipped upside down to create another (different) circular permutation.\\
Thus, we count the circular permutations of $8$ objects and then divide by 2, to get
$7!/2 = 2520$ different bracelets can be created.
\end{example}


\begin{problem}(problem 5)
Jean has 6 keys to put on a key ring, but does not want to put the house key 
next to the office key.  In how many ways can Jean configure the key ring? $\; \answer{36}$
\end{problem}





\end{document}

