\documentclass[handout]{ximera}

\input{../preamble}

\pgfplotsset{compat=1.13}

\outcome{Create harmonic conjugates}

\title{3.5 Harmonic Functions}

\begin{document}

\begin{abstract}
We determine and create harmonic functions.
\end{abstract}

\maketitle


\begin{definition}
A function $u(x,y)$ is called {\bf harmonic} on a disk $D \subset \R^2$ if the second order partial derivatives
$u_{xx}$ and $u_{yy}$ exist on $D$ and satisfy the equation
\[
\Delta u = u_{xx} + u_{yy} = 0
\]
on $D$.
\end{definition}

\begin{example}[example 1]
The function $u(x,y) = 3x + 5y +1$ is harmonic in the entire plane, $\R^2$ since
\[
u_{xx} \quad \text{and} \quad u_{yy} = 0
\]
everywhere. Hence $\Delta u = 0$ on $\R^2$ and $u$ is harmonic everywhere.
\end{example}

\begin{problem}(problem 1)
Select all of the functions that are harmonic on $\R^2$?\\
\begin{selectAll}
\choice[correct]{$u(x,y) = 2x -3y$}
\choice[correct]{$u(x,y) = xy$}
\choice{$u(x,y) = x^2 + y^2$}
\choice[correct]{$u(x,y) = x^2 - y^2$}
\end{selectAll}
\end{problem}


\begin{example}[example 2]
Show that the function $u(x,y) = \ln\left(x^2 + y^2\right)$ is harmonic in the 
punctured plane, $\R^2 \backslash (0,0)$.\\
Using the chain rule gives
\[
u_x = \frac{2x}{x^2 + y^2}
\]
Next, the quotient rule gives
\[
u_{xx} = \frac{(x^2+y^2)(2)- (2x)(2x)}{\left(x^2 +y^2\right)^2} = \frac{2y^2-2x^2}{\left(x^2 +y^2\right)^2}
\]
Due to the symmetry in variables in the formula for $u(x,y)$, we can see that
\[
u_{yy} = \frac{2x^2-2y^2}{\left(x^2 +y^2\right)^2}
\]
Adding these gives
\[
\Delta u = u_{xx} + u_{yy} = 0
\]
since the numerators are opposites (and the denominators are equal). 
The above calculations are valid for all $(x,y)\in \R^2$ except $(0,0)$ (where $u$ is not defined).  
Hence $u$ is harmonic in the punctured plane, $R^2\backslash (0,0)$.
\end{example}


\begin{problem}(problem 2)
Show that the following functions are harmonic on the indicated set.\\
\begin{align*}
a) \;\; u(x,y) &= e^x\cos y \;\;\text{on} \;\; \C\\
u_y &= \answer{-e^x \sin y}\\
u_{yy} &= \answer{-e^x \cos y}\\
b) \;\; u(x,y) &= \tan^{-1}\left(\frac{y}{x}\right) \;\; \text{on} \;\; \C\backslash \{y-\text{axis}\}\\
u_x &= \answer{-\frac{y}{x^2+y^2}}\\
u_{xx} &= \answer{\frac{2xy}{(x^2+y^2)^2}}
\end{align*}
\begin{hint}
Recall $\frac{d}{dx} \tan^{-1} x = \frac{1}{1+x^2}$
\end{hint}
\begin{hint}
Simplify $u_x$ before calculating $u_{xx}$
\end{hint}

\end{problem}

Here is a video solution to problem 2b:\\
\begin{foldable}
\youtube{QQg2smo_WOY}
\end{foldable}

\begin{theorem}
If $f(z) = u(x,y) + iv(x,y)$ is analytic in a disk $D$, and if the second order 
partial derivatives of $u$ and $v$ exist and are continuous on $D$ then $u$ and $v$ are both harmonic on $D$.
\end{theorem}
\begin{proof}
We prove that $u$ is harmonic on $D$ and leave the proof of $v$ as an exercise.\\
By the Cauchy-Riemann equations, $u_y = -v_x$ on $D$.  Therefore, $u_{yy} = -v_{xy} = -v_{yx}$ where the 
second equality follows from the symmetry of mixed partial derivatives. Applying the Cauchy-Riemann 
equations (specifically $v_y = u_x$)  gives
\[
u_{yy} = -v_{yx} = -u_{xx}
\]
on $D$. Hence
\[
\Delta u = u_{xx} + u_{yy} = u_{xx} - u_{xx} = 0
\]
and $u(x,y)$ is harmonic on $D$.
\end{proof}

\begin{problem}(harmonic)
Suppose $f(z) = u(x,y) + iv(x,y)$ satisfies the hypotheses of the above theorem on a disk $D$. 
Prove $v$ is harmonic on $D$.\\
To begin, note that $v$ is the real part of which function?
\begin{multipleChoice}
\choice{$-f(z)$}
\choice{$i f(z)$}
\choice[correct]{$-i f(z)$}
\end{multipleChoice}
Is this function analytic on $D$? \wordChoice{\choice[correct]{yes}\choice{no}}\\
How do we know $v$ is harmonic on $D$? \wordChoice{\choice{direct computation of $\Delta v$}\choice[correct]{the theorem above}}
\end{problem}
  
  
\begin{theorem}
Suppose $u(x,y)$ is harmonic on a disk $D$.  Then there exists a complex function $f(z)$, analytic on $D$, 
such that $u(x,y)$ is the real part of $f$. The imaginary part of $f$ is called the {\bf harmonic conjugate} of $u$.
\end{theorem}


\begin{example}[example 3]
Find the harmonic conjugate of $u(x,y) = x^3 - 3xy^2$.\\
First, we should verify that $u$ is harmonic:
\[
\Delta u = u_{xx} + u_{yy} = (6x) + (-6x) = 0
\]
so $u$ is harmonic on $\C$. According to the theorem we can find $v(x,y)$ such that
$f(z) = u(x,y) + iv(x,y)$ is analytic (entire in this case). To construct $v$, we begin by noting that
$v_y = u_x = 3x^2 - 3y^2$ since $u$ and $v$ must satisfy the Cauchy-Riemann equations.
Now, we integrate with respect to $y$ treating $x$ as a constant:
\[
v(x,y) = \int \left(3x^2 -3y^2\right) \; dy
\]
which yields
\[
v(x,y) = 3x^2y - y^3 + C(x)
\]
Note that the constant of integration is written as $C(x)$ since we were assuming in 
the integral that $x$ was constant. To determine $C(x)$, we use the other Cauchy-Riemann equation. 
On one hand, $v_x = -u_y = 6xy$ and on the other hand, since we have established a form for $v$, we can use it
to obtain $v_x = 6xy + C'(x)$. Equating these two expressions for $v_x$ we see 
that $C'(x) =0$ and hence $C(x)$ is a constant. Thus the harmonic conjugate of $u(x,y) = x^3 - 3xy^2$ is
\[
v(x,y) = 3x^2y - y^3 + C
\]
where $C$ is any real constant.
\end{example}

\begin{remark}
In general, a harmonic function $u$ has infinitely many harmonic conjugates, each pair differing by a constant.
\end{remark}

\begin{problem}(problem 3)
Verify that each function is harmonic on $\C$ and then find its harmonic conjugate.
\begin{align*}
a) \;\;u(x,y) &= x^4 -6x^2y^2 + y^4\\
v_y &= \answer{4x^3 - 12xy^2}\\
v(x,y) &= \answer{4x^3y - 4xy^3}+C\\[8pt]
b) \;\;u(x,y) &= e^x\sin y\\
v_y &= \answer{e^x \sin y}\\
v(x,y) &= \answer{-e^x \cos y}+C\\[8pt]
c) \;\;u(x,y) &= \cos x \cosh y\\
v_y &= \answer{-\sin x \cosh y}\\
v(x,y) &= \answer{-\sin x \sinh y}+C\\[8pt]
d) \;\; u(x,y) &= \sinh x \sin y\\
v_y &= \answer{\cosh x \sin y}\\
v(x,y) &= \answer{-\cosh x \cos y}+C
\end{align*}
\end{problem}

Here is a video solution to problem 3d:\\
\begin{foldable}
\youtube{QaRO5fWul8c}
\end{foldable}

\end{document}
