\documentclass[handout]{ximera}

\input{../preamble}

\outcome{Determine the behavior of an infinite series}

\title{3.9 Root Test}

\begin{document}

\begin{abstract}
Use the Root Test to determine whether an infinite series converges or diverges.
\end{abstract}

\maketitle

\section{Root Test}

In this section we will determine whether an infinite series converges or diverges using the Root Test. 
The idea behind this test is to determine if the given series is behaving similarly to a geometric series.
For a geometric series, 
\[
\sum_{n=0}^\infty ar^n,
\]
the $n^{th}$ root of the $n^{th}$ term approaches $r$ as $n \to \infty$ i.e., 
\[
\lim_{n \to \infty} \sqrt[n]{\left|ar^n\right|} = \lim_{n \to \infty} \left|r\right| \cdot \left|a\right|^{1/n} = r\cdot a^0 = r.
\]

%since for any number $a \neq 0$, 
%\[
%\lim_{n \to \infty}  \left|\sqrt[n]{a}\right| = 1.
%\]


For a general series, if the limit of the $n^{th}$ root of the absolute value of the $n^{th}$ term
approaches a value $L$ as $n \to \infty$, then the series behaves like a geometric series with ratio equal to $L$.
We state this precisely in the following theorem.

\begin{theorem}[Root Test]
For the series
\[
\sum_{n=0}^\infty a_n,
\]
let
\[
L = \lim_{n \to \infty} \sqrt[n]{\left|{a_n}\right|}.
\]
The series converges (absolutely) if $L < 1$ and diverges if $L>1$. 
If $L = 1,$ then no conclusion can be made.

\end{theorem}






\begin{example}[example 1]
Determine whether the series converges or diverges:
\[
\sum_{n=1}^\infty \left(\frac{n+ 2}{3n+1}\right)^n
\]


By the Root Test,
\[
L =  \lim_{n \to \infty} \sqrt[n]{\left|\left(\frac{n+ 2}{3n+1}\right)^n\right|}=  \lim_{n \to \infty} \sqrt[n]{\left|\frac{n+ 2}{3n+1}\right|^n} = \lim_{n \to \infty} \frac{n+ 2}{3n+1} = \frac13.
\]
Since $L = 1/3 < 1$, the series converges.

\end{example}



\begin{problem}(problem 1)

Does the series converge or diverge:
\[
\sum_{n=0}^\infty (-1)^n\left(\frac{n^2 + 2n+ 4}{4n^2 + 5}\right)^n
\]
\[
L = \lim_{n \to \infty} \sqrt[n]{\left|{a_n}\right|}= \answer{1/4}
\]
The series \wordChoice{\choice[correct]{converges}\choice{diverges}}
\end{problem}

\begin{center}
\begin{foldable}
\unfoldable{Here is a detailed, lecture style video on the Root Test:}
\youtube{nCYX4g4B_ZI}
\end{foldable}
\end{center}





\end{document}


\begin{example} %example #15
Find $h'(x)$ if $h(x) = x^{\sin(x)}$.\\
We will use the fact that the exponential and logarithm functions are inverses,
\[e^{\ln(x)} = x,\]
and the exponent property of logarithms, 
\[\ln(x^n) = n\ln(x),\]
to rewrite $h(x)$.  We have 
\[h(x) = x^{\sin(x)} = e^{\ln(x^{\sin(x)})} = e^{\sin(x)\ln(x)}\]
and we can now compute $h'(x)$ using a combination of the chain rule and product rule.
We can write $h(x)$ as a composition, $f(g(x))$ with 
\[g(x) = \sin(x)\ln(x) \quad \text{and} \quad f(x) = e^x.\]
Then to find $g'(x)$ we us the product rule and we get $g'(x) = \frac{\sin(x)}{x} + \cos(x)\ln(x)$.
Next $f'(x) = e^x$ and 
hence $f'(g(x)) = e^{g(x)} = e^{\sin(x)\ln(x)} = x^{\sin(x)}$.
We can then conclude $h'(x) = f'(g(x))g'(x) = x^{\sin(x)} \left[ \frac{\sin(x)}{x} + \cos(x)\ln(x)\right]$.
\end{example}

%more question formats below













%\begin{verbatim}
\begin{question}
What is your favorite color?
\begin{multipleChoice}
\choice[correct]{Rainbow}
\choice{Blue}
\choice{Green}
\choice{Red}
\end{multipleChoice}
\begin{freeResponse}
Hello
\end{freeResponse}
\end{question}
%\end{verbatim}





\begin{question}
  Which one will you choose?
  \begin{multipleChoice}
    \choice[correct]{I'm correct.}
    \choice{I'm wrong.}
    \choice{I'm wrong too.}
  \end{multipleChoice}
\end{question}


\begin{question}
  Which one will you choose?
  \begin{selectAll}
    \choice[correct]{I'm correct.}
    \choice{I'm wrong.}
    \choice[correct]{I'm also correct.}
    \choice{I'm wrong too.}
  \end{selectAll}
\end{question}


\begin{freeResponse}
What is the chain rule used for?
\end{freeResponse}
