\documentclass[handout]{ximera}

\input{../preamble}

\outcome{In this section we define vectors in two and three dimensions and manipulate them algebraically.}

\title{1.2 Vectors}



\begin{document}

\begin{abstract}
In this section we define vectors in two and three dimensions and manipulate them algebraically.
\end{abstract}
 
\maketitle

A {\bf vector} is a quantity that has both magnitude and direction. Basic examples of vectors include a force acting on an object and the velocity of an object in motion.
Vectors are represented graphically by arrows.  The length of the arrow represents the magnitude of the vector and the arrow points in the direction of the vector.
Algebraically, vectors are represented in a manner very similar to points.  In two dimensions a vector is written $\avec{v} = \vec{v} = \vector{x,y}$ and in three dimensions we write 
$\vec{v} = \vector{x,y, z}$.  In this context, $x, y$ and $z$ are called the {\bf components} of the vector, $\vec{v}$. 
The graphical representation of these vectors (in standard position)
would be as arrows emanating from the origin (initial point) and terminating at the point $(x,y)$ or $(x,y,z)$ (the final point).
$\avec{v}$
\section{Vectors in $\R^2$}


\end{document}







\begin{image}
\begin{tikzpicture}
\draw[thick, ->] (0,0) -- (2,0) node[right]{$y$};
\draw[thick, ->] (0,0) -- (0,1.8) node[above]{$z$};
\draw[thick, ->] (0,0) -- (-1.2,-0.9) node[below, left]{$x$};
\draw[blue] (0.3,0.2) -- (1.7, 0.5) -- (1.7, 1.7) -- cycle;
\draw[thin, blue]  (1.6, 0.47) -- (1.6, .59) -- (1.7, .62);
\draw[fill, blue] (0.3,0.2) circle [radius=0.03] node[left, blue]{$(x_1, y_1, z_1)$};
\draw[fill, blue] (1.7,1.7) circle [radius=0.03] node[right, blue]{$(x_2, y_2, z_2)$};
\draw[fill, blue] (1.7,0.5) circle [radius=0.037] node[right, blue]{$(x_2, y_2, z_1)$};
\node[below, blue] at (1.3,0.4) {$d_1$}; %adjacent
\node[right, blue] at (1.9,1) {$d_2$}; %opposite 
%\node[right] at (4.1,2) {$\tan(\theta)$};
\node[above, blue] at (1,1.4) {$d$}; %hypotenuse
%\node[below] at (1.75,-0.6) {$d = \sqrt{(x_2 - x_1)^2 + |y_2 - y_1|^2}$};
%\node[above] at (1.75,3.6) {A right triangle with $\cos(\theta) = x$};
\end{tikzpicture}
\end{image}
