\documentclass[handout]{ximera}

\input{../preamble}

\outcome{Simplify algebraic expressions using the rules for exponents}

\title{0.1 Exponent Rules}

\begin{document}

\begin{abstract}
We explore the rules for exponents.
\end{abstract}



\maketitle

\begin{center}
\textbf{Understanding Exponential Expressions}
\end{center}

An \textbf{exponential expression} takes the form $a^b$, where $a$ is the \textbf{base}  and 
$b$ is the \textbf{exponent}.
For instance, $2^3, x^n$ and $e^x$ are exponential expressions. 
When we encounter the expression $a^b$, 
where $b$ is a positive integer, it signifies the act of multiplying $a$
by itself $b$ times. Visually, this can be expressed as:
\[
a^b = \underbrace{a \cdot a \cdot \ldots \cdot a}_\text{$b$ times} 
\]


\begin{problem}(problem 1)
Compute the value of each of the following exponential expressions (without a calculator):
\[
a) \;6^2 = \answer{36} \quad b) \;7^2 = \answer{49} \quad c) \;8^2 = \answer{64} \quad d) \;9^2 = \answer{81}
\]
\end{problem}

\begin{problem}(problem 2)
Compute the value of each of the following exponential expressions (without a calculator):
\[
a) \ 2^3 = \answer{8} \; \ b) \ 3^3 = \answer{27} \; \  c) \ 4^3 = \answer{64} \; \  d) \ 5^3 = \answer{125} 
 \; \  e) \ 10^3 = \answer{1000}
\]
\end{problem}

\begin{problem}(problem 3)
Compute the value of each of the following exponential expressions (without a calculator):
\[
a) \ 2^4 = \answer{16} \; \  b) \ 3^4 = \answer{81} \; \  c) \ 5^4 = \answer{625} \; \ d) \ 10^4 = \answer{10000}
\]
\end{problem}

\begin{problem}(problem 4)
Compute the value of each of the following exponential expressions (without a calculator):
\[
a) \;2^5 = \answer{32} \quad b) \;2^6 = \answer{64} \quad c) \;2^7 = \answer{128} \quad d) \;2^8 = \answer{256}
\]
\end{problem}

Note that $64 = 8^2 = 4^3 = 2^6$.\\

According to the order of operations, exponents are computed before other 
multiplications and divisions. Hence, if $f(x) = 5x^2$ then $f(3) = 5(3^2) = 5(9) = 45$.

\begin{problem}(problem 5)
Evaluate the function $f(x) = 4x^2$ as indicated (without a calculator):
\[
a) \ f(2) = \answer{16}  \; \  b) \ f(-2) = \answer{16}  \; \  c) \ f(-3) = \answer{36}  \; \  d) \ f(-5) = \answer{100}
\]
\end{problem}

\begin{problem}(problem 6)
Compute the value of each of the following exponential expressions (without a calculator):
\[
a) \ -2^2 = \answer{-4}  \; \  b) \ (-2)^2 = \answer{4}  \; \  c) \ -1^4 = \answer{-1}  \; \  d) \ (-1)^4 = \answer{1}
\]
\end{problem}

\begin{center}
\textbf{Adding Exponents}
\end{center}
When multiplying exponential expressions with the same base, we add the exponents:
\[
a^b \cdot a^c = a^{b+c}
\] 
For positive integer exponents, the rationale is shown below:
\[
a^b \cdot a^c = (\underbrace{a \cdot a \cdot a \cdot \ldots \cdot a}_\text{$b$ times}) \cdot
(\underbrace{a \cdot a \cdot a \cdot \ldots \cdot a}_\text{$c$ times})
\]

\[
= \underbrace{a \cdot a \cdot a \cdot \ldots \cdot a}_\text{$b+c$ times}
\]

\[
= a^{b+c}
\]

\begin{problem}(problem 7)
Combine each of the following exponential expressions into a single exponential expression:
\[
a) \ a^4\cdot a^5 = \answer{a^9}  \; \  b) \ r^2 \cdot r^3  = \answer{r^5}  \; \  c) \ n\cdot n^6 = \answer{n^7} 
 \; \ d) \ x^{14}\cdot x^{17} = \answer{x^{31}}
\]
\end{problem}


\begin{problem}(problem 8)
Distribute: $x^2(x^4 + x^7)  = \answer{x^9+x^6}$
\end{problem}


\begin{center}
\textbf{Subtracting Exponents}
\end{center}
When dividing exponential expressions with the same base, we subtract the exponents:
\[
\frac{a^b}{a^c} = a^{b-c}
\] 
For positive integer exponents $b$ and $c$ with $b > c$, the rationale is shown below:
\begin{eqnarray*}
\frac{a^b}{a^c} &=& \frac{\overbrace{a \cdot a \cdot a \cdot \ldots \cdot a}^\text{$b$ times}}
{\underbrace{ a \cdot a \cdot \ldots \cdot a}_\text{$c$ times}}\\[10pt]
&=& \underbrace{ a \cdot a \cdot \ldots \cdot a}_\text{$b-c$ times}\\[10pt]
&=& a^{b-c}
\end{eqnarray*}

\begin{problem}(problem 9)
Combine each of the following exponential expressions into a single exponential expression:
\[
a) \ \frac{a^2}{a} = \answer{a}  \; \  b) \ \frac{r^3}{r} = \answer{r^2}  \; \  c) \ \frac{n^5}{n^3} = \answer{n^2} 
 \; \  d) \ \frac{x^7}{x^3} = \answer{x^4}
\]
\end{problem}


\begin{problem}(problem 10)
Divide: $\frac{x^8 + x^5 + x^2}{x^2} = \answer{x^6 + x^3 + 1}$
\end{problem}


\begin{center}
\textbf{The Zero Exponent}
\end{center}
For equal bases $a$,
\[
\frac{a^b}{a^b} = 1
\] 
Also, using the exponent subtraction rule, 
\[
\frac{a^b}{a^b} = a^{b-b} = a^0
\] 
Therefore, we define $a^0 = 1$ for $a \neq 0$.

\begin{definition} 
If $a \neq 0$, then we define $a^0 \equiv 1$.
\end{definition}

\begin{problem}(problem 11)
Evaluate each of the following exponential expressions (without a calculator):
\[
a) \ -3^0 = \answer{-1}  \; \  b) \ e^0 = \answer{1}  \; \  c) \ \pi^0 = \answer{1}  \; \  d) \ (-1)^0 = \answer{1}
\]
\end{problem}

\begin{center}
\textbf{Negative Exponents}
\end{center}
If the exponent in the denominator is greater than the exponent in the numerator, 
then subtracting exponents yields a negative exponent. Consider the following example:

\[
\frac{a^3}{a^5} = a^{3-5} = a^{-2}
\]
On the other hand, we can rewrite this ratio of exponential expressions as follows
\[
\frac{a^3}{a^5} = \frac{a\cdot a\cdot a}{a\cdot a\cdot a \cdot a\cdot a} = \frac{1}{a\cdot a} = \frac{1}{a^2}
\]
Putting these together yields
\[
a^{-2} = \frac{1}{a^2}
\]
We generalize this idea and define negative exponents as follows:
\begin{definition} For $a \neq 0$ and $n$ a positive integer, we define $a^{-n}$ by
\[
a^{-n} = \frac{1}{a^n}
\]
\end{definition}

\begin{problem}(problem 12)
Rewrite the following expressions using negative exponents:
\[
a) \ \frac{3}{x^5} = \answer{3x^{-5}}  \; \ b) \ \frac{1}{3x^5} = \answer{x^{-5}/3}  \; \ c) \ \frac{1}{u^2} = \answer{u^{-2}}
\]
\end{problem}

\begin{problem}(problem 13)
Simply the following expression using negative exponents:
\[
\frac{x^4 + x^2 + 1}{x^5} = \answer{x^{-1} + x^{-3} + x^{-5}}
\]
\end{problem}


\begin{center}
\textbf{Multiplying Exponents}
\end{center}

When raising an exponential expression to a power, we multiply the exponents. 
In the justification presented is below, the exponents, $b$ and $c$, are positive integers:
\[
\left(a^b\right)^c = \underbrace{a^b \cdot a^b  \cdot \ldots \cdot a^b}_\text{$c$ times}
\]
Adding the exponents gives:
\[
= a^{\overbrace{b + b + \dots +b}^\text{$c$ times}} = a^{bc}
\]

\begin{problem}(problem 14)
Combine each of the following into a single exponential expression (without a calculator):
\[
a) \ \left(a^4\right)^5 = \answer{a^{20}}  \; \  b) \ \left(r^2\right)^3  = \answer{r^6}  \; \  c) \ \left(n^6\right)^6 = \answer{n^{36}} 
 \; \  d) \ \left(x^{12}\right)^{12} = \answer{x^{144}}
\]
\end{problem}

\begin{center}
\textbf{Rational Exponents}
\end{center}
Let $n$ be a positive integer and let $a\geq 0$ if $n$ is even. 
Then  $\sqrt[n] a$ is the number which when multiplied by itself $n$ times gives $a$.
In other words:
\[
(\sqrt[n] a)^n = a
\]
Allow for a moment the use of the exponent $\frac{1}{n}$.  Using the rule for multiplying exponents, we have
\[
\left(a^\frac{1}{n}\right)^n = a^{\left(\frac{1}{n} \cdot n\right)} = a^1 = a
\]
Thus, $\sqrt[n] a$ and $a^\frac{1}{n}$ both yield $a$ when raised to the $n^\text{th}$ power.
Hence, we define $a^{1/n} = \sqrt[n] a$.
\begin{definition}
Let $n$ be a positive integer. We define
\[
a^{\frac{1}{n}} \equiv \sqrt[n] a
\]
where $a \geq 0 $ if $n$ is even.
\end{definition}

Now consider
\[
(\sqrt[n] a)^m = \left(a^{1/n}\right)^m = a^{m/n}
\]
This leads to the following definition.
\begin{definition}
Let $m$ and $n$ be positive integers. Then we define
\[
a^{\frac{m}{n}} \equiv (\sqrt[n] a)^m = \sqrt[n]{a^m}
\]
where $a \geq 0 $ if $n$ is even.
\end{definition}


\begin{problem}(problem 15)
Rewrite the following expressions using rational (and possibly negative) exponents:
\[
a) \ \frac{1}{\sqrt a} = \answer{a^{-1/2}}  \; \  b) \;\sqrt[3] r = \answer{r^{1/3}}  \; \  c) \ \sqrt[4]{u^3} = \answer{u^{3/4}}
\]
\end{problem}


\begin{problem}(problem 16)
Simplify into a sum of powers of $x$: 
\[
\frac{x^5 + \sqrt[3]{x^2} + x^{-3}}{x\sqrt x} = \answer{x^{7/2} + x^{-5/6} + x^{-9/2}}
\]
\end{problem}


\end{document}



