\documentclass{ximera}

\input{../preamble}

\outcome{Integrate rational functions using a partial fraction decomposition}

\title{Partial Fractions}

\begin{document}

\begin{abstract}
We will integrate rational functions using partial fraction decompositions.
\end{abstract}

\maketitle

\begin{center}
\textbf{Partial Fractions}
\end{center}

In this section we will learn an important method for integrating rational functions.
Recall that a rational function is a ratio of polynomial functions. The key background skill is the ability to factor polynomials.
We will look at three examples that cover the scenarios that we will encounter in our integrals. 

First, consider the cubic polynomial  $x^3 - x$.  
We can factor out an $x$ and then factor again using the difference of two perfect squares:
\[
x^3 - x = x(x^2-1) = x(x+1)(x-1).
\]

This is a \textbf{complete factorization} as there are three linear factors, $x, x-1$ and $x+1$, none of which has a constant factor.
We can say that the original cubic polynomial factored into three \textbf{distinct linear factors}.

Second, consider the cubic polynomial $x^3 + 2x^2 + x$.  This polynomial factors as $x(x^2 + 2x + 1)$ but this is not a complete factorization since
the quadratic $x^2 + 2x + 1$ is a perfect square we can be factored further.  The complete factorization is 
\[
x^3 + 2x^2 + x= x(x^2 + 2x + 1) = x(x+1)(x+1) = x(x+1)^2.
\]
In this case, our original cubic polynomial factored into a linear factor and a \textbf{repeated linear factor}.

Last, consider the cubic polynomial $x^3 + x^2$.  This polynomial factors as $x(x^2 + 1)$.
The quadratic $x^2 + 1$ does not factor (unless you are willing to 
use \link[imaginary numbers]{https://en.wikipedia.org/wiki/Imaginary_number}) and is called 
an \textbf{irreducible quadratic} polynomial.
So we say that the original cubic polynomial factored into a linear factor and an irreducible quadratic factor.



Now we will investigate the three decomposition forms we will encounter in our integrals.

\begin{itemize}

\item{\textbf{Distinct Linear Factors}} A rational function whose denominator factors into a product of distinct linear factors
can be decomposed into a sum of rational functions whose numerators are constants and whose denominators are the factors of the denominator.
For example 
\[
\frac{x^2 + 3x -5}{x^3 -x} = \frac{x^2 + 3x -5}{x(x-1)(x+1)} = \frac{A}{x} + \frac{B}{x-1} + \frac{C}{x+1},
\]
where $A, B$ and $C$ are constants.




\item{\textbf{Repeated Linear Factor}} The repeated linear factor will contribute a number of terms to the partial fraction 
decomposition equal to the \textbf{multiplicity} of the factor. Each of these terms will have a constant in the numerator and a 
distinct power of the repeated factor in the denominator.
For example 
\[
\frac{x^2 + 3x -5}{x^3 + x^2} = \frac{x^2 + 3x -5}{x^2(x+1)} = \frac{A}{x} + \frac{B}{x^2} + \frac{C}{x+1},
\]
where $A, B$ and $C$ are constants. Note how the repeated linear factor $x^2$ contributed two terms whose denominators were $x$ and $x^2$
and whose numerators were constants.


\item{\textbf{Irreducible Quadratic Factor}} An irreducible quadratic factor in the denominator will contribute one term to the decomposition whose numerator
is a linear polynomial and whose denomiator is the irreducible quadratic factor.
For example 
\[
\frac{x^2 + 3x -5}{x^3 +x} = \frac{x^2 + 3x -5}{x(x^2+1)} = \frac{A}{x} + \frac{Bx + C}{x^2 + 1},
\]
where $A, B$ and $C$ are constants.


\end{itemize}

Before we move on to integrating the special rational functions seen in the above decompositions, 
we will look at one more example consisting of all three of the ideas mentioned above:
 
\begin{example} %example #1
Find the partial fraction decomposition of
\[
  f(x) = \frac{x^2 + 3x -5}{x(x-2)^3(x^2 + 4)}. 
\]
The denominator is already in a completely factored form.  This form consists of a linear factor $x$, 
a repeated linear factor $(x-2)^3$ and an irreducible quadratic $x^2 + 4$. 
The partial fraction decomposition will consist of one term for the factor $x$ and three terms for the factor
$(x-2)^3$.  Each of these terms will have a constant in the numerator.  
The decomposition will also contain one term for the irreducible quadratic and this term 
will have a linear polynomial in the numerator and the irreducible quadratic in the denominator.
The form of the decomposition is:
\[
  \frac{x^2 + 3x -5}{x(x-2)^3(x^2 + 4)} = \frac{A}{x} + \frac{B}{x-2} + \frac{C}{(x-2)^2} + \frac{D}{(x-2)^3} +\frac{Ex+F}{x^2+4},
\]
where $A, B, C, D, E$ and $F$ are constants.

\end{example}

\begin{center}
\begin{foldable}
\unfoldable{Here is a video of Example 1}
%\youtube{Yy6QXnFlnXs} %vid of example 1
\end{foldable}
\end{center}


\begin{problem} %problem #1
  Find the average...
  \[
  f(x)
  \]
    \begin{hint}
      compute a definite integral
    \end{hint}
    
		
		The average value is
		 $\answer[given]{0}$
\end{problem}


\begin{center}
\begin{foldable}
\unfoldable{Here is a detailed, lecture style video on average value:}
\youtube{gRJTe1i1muo}
\end{foldable}
\end{center}





\end{document}


\begin{example} %example #15
Find $h'(x)$ if $h(x) = x^{\sin(x)}$.\\
We will use the fact that the exponential and logarithm functions are inverses,
\[e^{\ln(x)} = x,\]
and the exponent property of logarithms, 
\[\ln(x^n) = n\ln(x),\]
to rewrite $h(x)$.  We have 
\[h(x) = x^{\sin(x)} = e^{\ln(x^{\sin(x)})} = e^{\sin(x)\ln(x)}\]
and we can now compute $h'(x)$ using a combination of the chain rule and product rule.
We can write $h(x)$ as a composition, $f(g(x))$ with 
\[g(x) = \sin(x)\ln(x) \quad \text{and} \quad f(x) = e^x.\]
Then to find $g'(x)$ we us the product rule and we get $g'(x) = \frac{\sin(x)}{x} + \cos(x)\ln(x)$.
Next $f'(x) = e^x$ and 
hence $f'(g(x)) = e^{g(x)} = e^{\sin(x)\ln(x)} = x^{\sin(x)}$.
We can then conclude $h'(x) = f'(g(x))g'(x) = x^{\sin(x)} \left[ \frac{\sin(x)}{x} + \cos(x)\ln(x)\right]$.
\end{example}

%more question formats below













%\begin{verbatim}
\begin{question}
What is your favorite color?
\begin{multipleChoice}
\choice[correct]{Rainbow}
\choice{Blue}
\choice{Green}
\choice{Red}
\end{multipleChoice}
\begin{freeResponse}
Hello
\end{freeResponse}
\end{question}
%\end{verbatim}





\begin{question}
  Which one will you choose?
  \begin{multipleChoice}
    \choice[correct]{I'm correct.}
    \choice{I'm wrong.}
    \choice{I'm wrong too.}
  \end{multipleChoice}
\end{question}


\begin{question}
  Which one will you choose?
  \begin{selectAll}
    \choice[correct]{I'm correct.}
    \choice{I'm wrong.}
    \choice[correct]{I'm also correct.}
    \choice{I'm wrong too.}
  \end{selectAll}
\end{question}


\begin{freeResponse}
What is the chain rule used for?
\end{freeResponse}
