\documentclass[handout]{ximera}

\input{../preamble}

\pgfplotsset{compat=1.13}

\outcome{Determine the radius of convergence of complex power series}

\title{4.3 Complex Power Series}

\begin{document}

\begin{abstract}
We determine radius of convergence of complex power series.
\end{abstract}

\maketitle


\begin{definition} 
A {\bf complex power series centered at $z_0$} is a complex series of the form
\[
\sum_{k=0}^\infty c_k(z-z_0)^k = c_0 + c_1 (z-z_0) + c_2 (z-z_0)^2 + c_3 (z-z_0)^3 + \cdots
\]
where $z_0$ is a given complex number and $z$ is a complex variable.
\end{definition}
Since a complex power series contains a variable, the question is, for which values of the variable, $z$, 
does the complex power series converge? This question is answered by the root test as described in the following theorem.

\begin{theorem}
The complex power series centered at $z_0$, $\dis \sum_{k=0}^\infty c_k(z-z_0)^k$ will converge for 
$z \in D(z_0, \rho)$ and diverge for $|z-z_0| > \rho$, where $\rho = \frac{1}{L}$ with 
\[
L = \limsup_{k \to 0} \sqrt[k]{|c_k|}
\]
The radius, $\rho$ is called the radius of convergence of the complex power series.
If $L = 0$, then $\rho$ is understood to be infinity, i.e., the complex power series converges for all $z$.
If $L = \infty$, then the power series converges only at its center, $z_0$.
A complex power series may converge or diverge points on the circle $C(z_0, \rho)$.
\end{theorem}
\begin{remark}
The limit $L$, can also be computed from the ratio test as 
\[
L = \limsup_{k \to \infty} \left|\frac{c_{k+1}}{c_k}\right|
\]
provided $c_k \neq 0$ for all but finitely many values of $k$
\end{remark}


\begin{proof}
From the root test, the series  $\dis \sum_{k=0}^\infty c_k(z-z_0)^k$ converges if $z$ satisfies the inequality
\[
\limsup_{k \to 0} \sqrt[k]{|c_k(z-z_0)^k|} = |z-z_0| \limsup_{k \to 0} \sqrt[k]{|c_k|} = |z-z_0| \cdot L < 1
\]
If $L \neq 0$, then the series converges if $|z-z_0| < \frac{1}{L} = \rho$ and if $L =0$, then the series converges for all $z$.
Moreover, the same calculation also shows that if $|z-z_0| \cdot L > 1$ the series diverges.
\end{proof}

\begin{example}[example 1]
Find the disk of convergence of the complex power series: $\dis \sum_{k=1}^\infty \frac{i^k}{k}(z+i)^k$\\
We first compute the limit superior from the root test:
\[
L = \limsup_{k \to \infty} \sqrt[k]{|c_k|} = \limsup_{k \to \infty} \sqrt[k]{|\frac{i^k}{k}|} = \limsup_{k \to \infty} \frac{1}{\sqrt[k]{k}} = 1
\]
Therefore the radius of convergence is $\rho = 1/L = 1$ and since the center of the power series is $-i$, the disk of convergence is $D(-i, 1)$.
\end{example}

\begin{problem}(problem 1)
Find the disk of convergence of the complex power series: $\dis \sum_{k=1}^\infty \frac{e^{ik}}{k^2}(z-2i)^k$\\
The center of the power series is $\; \answer{2i}$\\
The limit superior of the coefficients from the root test (or ratio test) is:
\[
\limsup_{k \to \infty} \sqrt[k]{|c_k|} = \; \answer{1}
\]
The disk of convergence is:
\begin{multipleChoice}
\choice{$D(-2i, 1)$}\\
\choice{$D(2i, -1)$}\\
\choice[correct]{$D(2i, 1)$}\\
\end{multipleChoice}
\end{problem}

On its disk of convergence, a power series is an analytic function.
\begin{theorem}
Suppose the complex power series
\[
f(z) = \sum_{k=0}^\infty c_k(z-z_0)^k
\]
converges on $D(z_0, \rho)$ for some $\rho > 0$.
Then $f$ is differentiable on $D(z_0, \rho)$ and 
\[
f'(z) = \sum_{k=1}^\infty kc_k(z-z_0)^{k-1}
\]
Moreover, $f'$ has the same disk of convergence as $f$.
\end{theorem}
\begin{remark}
The formula for $f'$ shows that a power series can be differentiated term by term.
\end{remark}
\begin{corollary}
Suppose the complex power series
\[
f(z) = \sum_{k=0}^\infty c_k(z-z_0)^k
\]
converges on $D(z_0, \rho)$ for some $\rho > 0$.
Then $f$ is infinitely differentiable on $D(z_0, \rho)$
\end{corollary}

\begin{proof}
Since $f'$ is a power series with the same radius of convergence as $f$, the theorem can be applied to $f'$ and hence
$f'$ is differentiable and $f''$ has the same radius of convergence as $f'$ and hence, as $f$. Continuing inductively, $f$
has derivatives of al orders, i.e., $f$ is infinitely differentiable on $D(z_0, \rho)$.
\end{proof}

\begin{corollary}
Suppose the complex power series
\[
f(z) = \sum_{k=0}^\infty c_k(z-z_0)^k
\]
converges on $D(z_0, \rho)$ for some $\rho > 0$. Then we have the following formula for the coefficients:
\[
c_k = \frac{f(k)(z_0)}{k!}
\]
\end{corollary}
\begin{proof}
Since $f$ is infinitely differentiable and the derivative is computed term by term, we have
\[
f^{(j)}(z) = \sum_{k=j}^\infty k(k-1)(k-2)\cdots(k-j+1)c_k(z-z_0)^{k-j}
\]
Plugging in $z_0$ for $z$ and noting that all of the terms are zero except the first (where $k = j$),
we get
\[
f^{(j)}(z_0) = j!c_j
\]
Note that we treat the zero power in the first term as 1 when $z=z_0$.
\end{proof}
\end{document}



\begin{theorem}[Test for Divergence]
If $\dis \sum_{k=1}^\infty c_k$ converges, then $\dis \lim_{k\to\infty} c_k = 0$
\end{theorem}
\begin{proof}
Note that $c_n = S_n - S_{n-1}$, the difference in partial sums. Since the sequence $\{S_n\}$ converges, given $\epsilon >0$,
there exists a number $N$ and a number $S$ such that if $n>N$ then $|S_n - S| < \epsilon/2$. Hence for $n > N+1$, the triangle inequality gives
\begin{align*}
|c_n| &= |S_n - S_{n-1}| = |(S_n -S) + (S-S_{n-1})| \\
&\leq |S_n -S| + |S-S_{n-1}| < \frac{\epsilon}{2} + \frac{\epsilon}{2} =\epsilon
\end{align*}
Hence, the sequence $\{c_n\}$ converges to zero.
\end{proof}


\begin{example}[example 1] 
Determine whether the complex series $\dis \sum_{k=1}^\infty \frac{1+ik}{k}$ converges.\\
Since, $\dis \lim_{k \to \infty} \frac{1+ik}{k} = i$. Since $i \neq 0$, the Test for Divergence says that the series $\dis \sum_{k=1}^\infty \frac{1+ik}{k}$
does not converge, i.e., it diverges.
\end{example}

\begin{problem}(problem 1) 
Determine whether the complex series $\dis \sum_{k=0}^\infty \frac{k!}{(2i)^k}$ converges.\\
Since, $\dis \lim_{k \to \infty} \left|\frac{k!}{(2i)^k}\right| = \answer{\infty}$,\\
 $\dis \lim_{k \to \infty} \frac{k!}{(2i)^k} \neq \answer{0}$,\\
 the series $\dis \sum_{k=0}^\infty \frac{k!}{(2i)^k}\;\;$ \wordChoice{\choice{converges}\choice[correct]{diverges}}\\
 by the \wordChoice{\choice{Test for Convergence}\choice[correct]{Test for Divergence}}
\end{problem}

\begin{theorem}[Geometric Series]
Let $c \in \C$ and consider the series
\[
\sum_{k=0}^\infty c^k = 1 + c + c^2 + c^3 + \cdots
\]
This series converges if and only if $|c| < 1$.  Moreover, if it converges, its sum is 
$\displaystyle \frac{1}{1-c}$.
\end{theorem}

\begin{remark}
The complex number $c$ in the geometric series $\dis \sum_{k=0}^\infty c^k$ is called the common ratio of the series.
\end{remark}

\begin{proof}
Suppose $|c| < 1$. Then the partial sums have a closed form whose limit can be computed directly.
Since
\[
S_n = \sum_{k=1}^n c^k = 1+c+c^2+\cdots + c^n
\]
we have
\[
cS_n = c+c^2+\cdots + c^{n+1}
\]
Subtracting gives
\[
S_n - cS_n = 1-c^{n+1}
\]
from which we obtain the closed form
\[
S_n = \frac{1-c^{n+1}}{1-c}
\]
Now, since $|c|<1$, we have $c^n \to 0$ as $n \to \infty$ and hence
\[
\sum_{k=0}^\infty c^k = \lim_{n\to \infty} S_n = \lim_{n\to \infty} \frac{1-c^{n+1}}{1-c} = \frac{1}{1-c}
\]
Thus, the geometric series converges and its sum is also known.\\
If $|c| \geq 1$ then $c^n \nrightarrow 0$ as $n \to \infty$ which implies 
that the geometric series diverges (by the Test for Divergence).
\end{proof}


\begin{example}[example 2] 
Determine whether the geometric series converges or diverges:\\
a) $\;\;\dis \sum_{k=0}^\infty \frac{1}{(1+i)^k}\quad$  b) $\;\;\dis \sum_{k=0}^\infty \left(\frac{1+i}{1-i}\right)^k$ \\
Part a) This is a geometric series with common ratio $\dis \frac{1}{1+i}$. The modulus of the common ratio is 
\[
\left|\frac{1}{1+i}\right| = \frac{1}{\sqrt 2} <1
\]
Hence the geometric series converges and its sum is
\[
\sum_{k=0}^\infty \frac{1}{(1+i)^k} = \frac{1}{1-\frac{1}{1+i}} = \frac{1+i}{i} = 1-i
\]
Part b) This is a geometric series with common ratio $\dis \frac{1+i}{1-i}$. The modulus of the common ratio is 
\[
\left|\frac{1+i}{1-i}\right| = \frac{\sqrt 2}{\sqrt 2} = 1 \nless 1
\]
Hence the geometric series diverges.
\end{example}


\begin{problem}(problem 2)
Determine whether the geometric series converges or diverges. If it converges, find its sum: $\;\;\dis \sum_{k=0}^\infty \left(\frac{2i}{2+i}\right)$ \\
The common ratio is $\; \answer{\frac{2i}{2+i}}$\\
The modulus of the common ratio is $\; \answer{\frac{2}{\sqrt 5}}$\\
Is this less than 1? \wordChoice{\choice[correct]{yes}\choice{no}}\\
The geometric series  \wordChoice{\choice[correct]{converges}\choice{diverges}}\\
The sum of the series is $\; \answer{\frac{2+i}{2-i}}$
\end{problem}


\begin{theorem}[Absolute Convergence]
If the series $\dis \sum_{k=1}^\infty |c_k|$ converges, then so does the complex series $\dis \sum_{k=1}^\infty c_k$.
\end{theorem}


\begin{remark}
To prove this result, we will need the Monotone Convergence Theorem which states that a bounded, monotone sequence of real numbers converges.
\end{remark}

\begin{proof}
Suppose $\displaystyle \sum_{k=1}^\infty |c_k|$ converges to the real number $S$, and break the partial sums of  the complex series 
$\displaystyle \sum_{k=1}^\infty c_k$ into their real and imaginary parts:
\[
\sum_{k=1}^n c_k =  \sum_{k=1}^n a_k + i\sum_{k=1}^n b_k
\]
where $a_k, b_k \in \R$ for all $k$.  Let $a_k^+ = a_k$ if $a_k \geq 0$ and let $a_k^+ = 0$ if $a_k < 0$.
Similarly, let $a_k^- = -a_k$ if $a_k < 0$ and let $a_k^- = 0$ if $a_k \geq 0$. Then 
\[
\sum_{k=1}^n a_k = \sum_{k=1}^n a_k^+ - \sum_{k=1}^n a_k^- = A_n^+ - A_n^-
\]
The sequences $\{A_n^+\}$ and $\{A_n^-\}$ are both increasing sequences and since $|a_k| \leq |c_k|$ for all $k$, they are both bounded above by $S$.
Hence, by the Monotone Convergence Theorem, both sequences converge.
Thus, the real series $\dis \sum_{k=1}^n a_k$ converges, to a sum, $A$. Applying the same argument to the partial sums of the imaginary part, 
$\dis \sum_{k=1}^n b_k$, we see that it too converges to a sum, $B$. Finally, we can conclude that $\dis \sum_{k=1}^\infty c_k$ converges to $A + iB$. 
\end{proof}

\begin{example}[example 3]
Show that the complex series converges: $\dis \sum_{k=1}^\infty \frac{2e^i}{(k+ik)^2}$\\
The modulus of $c_k$ is
\[
\left|\frac{2e^i}{(k+ik)^2}\right| = \frac{2}{2k^2} = \frac{1}{k^2}
\]
and the series
\[
\sum_{k=1}^\infty \frac{1}{k^2}
\]
is a convergent $p$-series with $p = 2 > 1$. Moreover the sum of this well known series is $\frac{\pi^2}{6}$.
\end{example}

\begin{problem}(problem 3) 
Determine whether the complex series $\dis \sum_{k=0}^\infty \frac{i^k}{k!}$ converges.\\
The modulus of $c_k$ is $\; \answer{1/k!}$\\
The series $\dis \sum_{k=0}^\infty \frac{1}{k!}\;\;$ \wordChoice{\choice[correct]{converges}\choice{diverges}}\\
The series $\dis \sum_{k=0}^\infty \frac{i^k}{k!}\;\;$ \wordChoice{\choice[correct]{converges}\choice{diverges}}
\end{problem}


\begin{definition}
We define the limit superior of a sequence of real numbers $\{a_n\}, \; \limsup_{n \to \infty} a_n,\;$to be the largest limit of any subsequence of $\{a_n\}$.\
\end{definition}


\begin{remark}
If the sequence $\{a_n\}$ converges to $a$, then $\limsup_{n \to \infty} a_n =a$.
\end{remark}

\begin{remark}
It is possible that $\limsup_{n \to \infty} a_n =\pm\infty$
\end{remark}

\begin{example}[example 4]
Find the limit superior of the sequence $\{c_k\}= \{1, 2, 3, 1, 2, 3, 1, 2, 3, \ldots\}$
The sequence $\{c_k\}$ has subsequences that converge to $1, 2$ and $3$. Since the largest of these limits is $3$, we have
\[
\limsup_{k\to\infty} c_k = 3
\]
\end{example}

\begin{problem}(problem 4)
Find the limit superior of the sequence:\\
a) $\dis\; \{(-1)^k\} \quad \limsup_{k\to\infty} (-1)^k = \; \answer{1}$\\
a) $\dis\; \{c_k\} = \{1, -1, \frac12, -1, \frac13, -1, \ldots\} \quad \limsup_{k\to\infty} c_k = \; \answer{0}$\\
\end{problem}


\begin{theorem}[Root Test]
Let $\displaystyle \sum_{k=1}^\infty c_k$ be a complex series and let 
\[
L = \limsup_{k \to \infty} \sqrt[k]{|c_k|}
\]
Then the series converges if $L <1$ and diverges if $L >1$.
\end{theorem}

\begin{proof}
First, suppose $\dis L = \limsup_{k \to \infty} \sqrt[k]{|c_k|} < 1$. Then there exists a natural number $N$ such that for $k>N 
|c_k| < L^k$.  Hence the series $\dis \sum_{k = N+1}^\infty |c_k|$ converges by direct comparison to the geometric series
$\dis \sum_{k = N+1}^\infty L^k$. Thus, the series $\dis\sum_{k = 1}^\infty c_k$ converges absolutely and is therefore convergent.\\
Next, suppose $\dis L = \limsup_{k \to \infty} \sqrt[k]{|c_k|} > 1$ Then there exists a subsequence of $\{c_k\}$ such that $\sqrt[k]{|c_k|} > 1$
which implies that $|c_k| > 1$ for each of the infinitely many terms in this subsequence. Hence, $c_k \nrightarrow 0$ as $k \to \infty$.
By the Test for Divergence, the series $\displaystyle \sum_{k=1}^\infty c_k$ diverges.
\end{proof}

\begin{remark}
The series $\dis \sum_{k=1}^\infty \frac{1}{k^2}$ and $\dis \sum_{k=1}^\infty \frac{1}{k}$ show that if $L = 1$ in the Root Test,
then no conclusion can be drawn. This is because the first series converges and the second series diverges but $L=1$ in both cases.
\end{remark}

\begin{example}[example 5]
Determine if the series converges or diverges: $\dis \sum_{k=1}^\infty \left(\frac{k+i}{k^2 - i}\right)^k$\\
Computing the $k^{th}$ root of the modulus of $c_k$ gives
\[
\sqrt[k]{|c_k|} = \sqrt[k]{\left|\left(\frac{k+i}{k^2 - i}\right)^k\right|} = \sqrt[k]{\left|\frac{k+i}{k^2 - i}\right|^k} = \sqrt{\frac{k^2 + 1}{k^4 + 1}}
\]
and the limit superior is
\[
\limsup_{k \to \infty} \sqrt[k]{|c_k|} = \limsup_{k \to \infty}  \sqrt{\frac{k^2 + 1}{k^4 + 1}} = 0 <1
\]
so the series converges.
\end{example}


\begin{problem}(problem 5)
Determine if the series converges or diverges: 
\[
\sum_{k=1}^\infty \left(\frac{e^{ik}}{1+i\sqrt k}\right)^k
\]
$\sqrt[k]{|c_k|} = \; \answer{\frac{1}{\sqrt{1+k}}}$\\
The limit superior is $\; \limsup_{k\to\infty}\sqrt[k]{|c_k|}= \; \answer{0}$\\
Is this less than 1? \wordChoice{\choice[correct]{yes}\choice{no}}\\
The series  \wordChoice{\choice[correct]{converges}\choice{diverges}}
\end{problem}

\begin{theorem}[Ratio Test]
Let $\displaystyle \sum_{k=1}^\infty c_k$ be a complex series and let 
\[
L = \limsup_{k \to \infty} \left|\frac{c_{k+1}}{c_k}\right|
\]
Then the series converges if $L <1$ and diverges if $L >1$.
\end{theorem}

\begin{remark} 
If $c_k = 0$ for infinitely many values of $k$, then the ratio test cannot be used.
\end{remark}
\begin{remark} 
The proof of the ratio test is similar to the proof of the root test.
\end{remark}

\begin{example}[example 6]
Determine if the series converges or diverges: 
\[
\sum_{k=0}^\infty \frac{i^k2^k}{k!}
\]
The modulus of the ratio of successive terms is
\[
\left|\frac{i^{k+1}2^{k+1}}{(k+1)!}\cdot \frac{k!}{i^k 2^k}\right| = \frac{2}{k+1}
\]
Taking the limit gives
\[
\limsup_{k \to \infty} \frac{2}{k+1} = 0 < 1
\]
By the Ratio Test, the series converges.
\end{example}

\begin{problem}(problem 6)
Determine if the series converges or diverges: 
\[
\sum_{k=0}^\infty \frac{2e^{ik}}{(2k)!}
\]
$\left|\frac{c_{k+1}}{c_k}\right| = \; \answer{\frac{1}{(k+1)(2k+1)}}$\\
The limit superior is $\; \limsup_{k\to\infty}\left|\frac{c_{k+1}}{c_k}\right|= \; \answer{0}$\\
Is this less than 1? \wordChoice{\choice[correct]{yes}\choice{no}}\\
The series  \wordChoice{\choice[correct]{converges}\choice{diverges}}
\end{problem}


\end{document}




