\documentclass[handout]{ximera}

\input{../preamble}

\outcome{Derive differentiation rules for vector-valued functions.}

\title{2.3 Differentiation Rules}



\begin{document}

\begin{abstract}
In this section we will derive differentiation rules for vector-valued functions.
\end{abstract}

\maketitle

\section{Sum and Difference Rules}


The derivative of the sum or difference of two vector-valued functions is the sum or difference of their derivatives.

\begin{proposition}[Sum and Difference Rules]
Let $\vec r_1(t)$ and $\vec r_2(t)$ be vector-valued functions. Then
\[
\frac{d}{dt}\left[\vec r_1(t) \pm \vec r_2(t) \right] = \vec r_1\,'(t) \pm \vec r_2\,'(t)
\]
\end{proposition}
The reader should verify that the previous proposition is a direct consequence of the 
componentwise sum of vectors and the analogous differentiation rules for ordinary functions.

\section{Product Rules}
There are three types of multiplication involving vectors: multiplication by a scalar, the dot product, and the cross product.
We will use the product rule for ordinary functions to derive a product rule for all three of these operations.
Recall the product rule for functions $f(x)$ and $g(x)$:
\[
\frac{d}{dx}\left[f(x)g(x)\right] = f(x)g'(x) + f'(x)g(x)
\]
We begin with scalar multiplication.

\begin{proposition}[Product Rule for Scalar Multiplication]
Let $\vec r(t) = \vector{x(t), y(t),z(t)}$ be a vector-valued function with differentiable components, and let $f(t)$ be an ordinary, differentiable real-valued function.
Then
\[
\frac{d}{dt}\left[f(t)\vec r(t)\right] = f(t)\vec r\,'(t)) + f'(t)\vec r(t)
\]
\end{proposition}

\begin{proof}
Multiplying $f(t)$ into each component of $\vec r(t)$ and then differentiating yields
\begin{align*}
\frac{d}{dt}\left[f(t)\vec r(t)\right] &= \frac{d}{dt} \left[ \vector{f(t)x(t), f(t)y(t),f(t)z(t)}\right]\\
&=  \vector{\frac{d}{dt} \left[f(t)x(t)\right], \frac{d}{dt} \left[f(t)y(t)\right],\frac{d}{dt} \left[f(t)z(t)\right]}\\
&= \vector{f(t)x'(t) + f'(t)x(t), f(t)y'(t) + f'(t)y(t), f(t)z'(t) + f'(t)z(t)}\\
&=\vector{f(t)x'(t) , f(t)y'(t), f(t)z'(t)}+ \vector{f'(t)x(t), f'(t)y(t), f'(t)z(t)}\\
&= f(t)\vector{x'(t), y'(t), z'(t)} + f'(t)\vector{x(t), y(t), z(t)}\\
&= f(t)\vec r\,'(t)) + f'(t)\vec r(t)
\end{align*}
as claimed.
\end{proof}

\begin{proposition}[Product Rule for the Dot Product]
Let $\vec r_1(t) = \vector{x_1(t), y_1(t),z_1(t)}$ and $\vec r_2(t) = \vector{x_2(t), y_2(t),z_2(t)}$ be vector-valued functions with differentiable components.
Then
\[
\frac{d}{dt}\left[\vec r_1(t) \dotp \vec r_2(t)\right] = \vec r_1(t)\dotp \vec r_2\,'(t) + \vec r_1\,'(t)\dotp\vec r_2(t)
\]
\end{proposition}
\begin{proof}
Computing the dot product and then differentiating yields
\begin{align*}
\frac{d}{dt}\left[\vec r_1(t) \dotp \vec r_2(t)\right] &= \frac{d}{dt} \left[ \vector{x_1(t)x_2(t), y_1(t)y_2(t), z_1(t)z_2(t)}\right]\\
&=  \vector{\frac{d}{dt} \left[x_1(t)x_2(t)\right], \frac{d}{dt} \left[y_1(t)y_2(t)\right],\frac{d}{dt} \left[z_1(t)z_2(t)\right]}\\
&= \vector{x_1(t)x_2'(t) +x_1'(t)x_2(t), y_1(t)y_2'(t) +y_1'(t)y_2(t), z_1(t)z_2'(t) +z_1'(t)z_2(t)}\\
&=\vector{x_1(t)x_2'(t) , y_1(t)y_2'(t), z_1(t)z_2'(t)}+ \vector{x_1'(t)x_2(t), y_1'(t)y_2(t), z_1'(t)z_2(t)}\\
&= \vector{x_1(t), y_1(t), z_1(t)}\dotp \vector{x_2'(t), y_2'(t), z_2'(t)} + \vector{x_1'(t), y_1'(t), z_1'(t)}\dotp \vector{x_2(t), y_2(t), z_2(t)}\\
&= \vec r_1(t) \dotp \vec r_2\,'(t) + \vec r_1\,'(t) \dotp \vec r_2(t)
\end{align*}
as claimed.
\end{proof}

\begin{proposition}
Let $\vec r(t)$ be a vector-valued function and suppose $|\vec r(t)|$ is constant. Then $\vec r(t) \dotp \vec r\,'(t) = 0$.\\
\end{proposition}
\begin{proof}
Since $|\vec r(t)|^2 = c$, a constant, we have
\[
\frac{d}{dt} |\vec r(t)|^2 = 0
\]
However, $|\vec r(t)|^2$ can be expressed using the dot product and computing this derivative using the product rule for the dot product gives,
\begin{align*}
0 &= \frac{d}{dt} |\vec r(t)|^2 \\
&= \frac{d}{dt} \left[\vec r(t) \dotp \vec r(t)\right] \\
  &= \vec r(t) \dotp \vec r\,'(t) + \vec r\,'(t) \dotp \vec r(t) \\
  &= 2 \left[\vec r(t) \dotp \vec r\,'(t)\right]
\end{align*}
The last equation is due to the fact that the dot product is commutative. Hence $\vec r(t) \dotp \vec r\,'(t) = 0$
\end{proof}

\begin{remark}
The geometric interpretation of the above proposition is that if the graph of space curve $\vec r(t)$ is contained 
on the surface of a sphere with center at the origin,
then its tangent vectors will be orthogonal to its position vectors.  
This is due to the fact that for such a sphere, the position vector of any point on the sphere is 
orthogonal to the plane tangent to the sphere at that point.
\end{remark}

\begin{problem}(Problem 1)
Let $\vec r(t)$ be a vector-valued function. Show that if $\vec r(t) \neq \vec 0$ then 
\[
\frac{d}{dt} |\vec r(t)| = \frac{1}{|\vec r(t)|} \vec r(t) \dotp \vec r\,'(t)
\]
\end{problem}

\begin{definition}[Unit Tangent Vector]
Let $\vec r(t)$ be a vector-valued function. Then the \bf{unit tangent vector}, $\vec t(t)$ is defined as
\[
\vec T(t) = \frac{\vec r\,'(t)}{|\vec r\,'(t)|}
\]
Note that $\vec T(t)$ is in fact a unit vector.
\end{definition}

\begin{corollary}
Let $\vec r(t)$ be a vector valued function and let $\vec T(t)$ be the unit tangent vector.
Then 
\[
\vec T(t) \dotp \vec T\,'(t) = 0
\]
\end{corollary}
\begin{proof}
The magnitude of the unit tangent vector, $\vec T(t)$, is 1 and hence constant. The result then follows from the proposition.
\end{proof}


\begin{proposition}[Product Rule for the Cross Product]
Let $\vec r_1(t) = \vector{x_1(t), y_1(t),z_1(t)}$ and $\vec r_2(t) = \vector{x_2(t), y_2(t),z_2(t)}$ be vector-valued functions with differentiable components.
Then
\[
\frac{d}{dt}\left[\vec r_1(t) \cross \vec r_2(t)\right] = \vec r_1(t) \cross \vec r_2\,'(t) + \vec r_1\,'(t) \cross \vec r_2(t)
\]
\end{proposition}
\begin{proof}
To simply the notation, we will suppress the variable and write $x_1$ and $x_1'$ in place of $x_1(t)$ and $x_1'(t)$ and similarly for $y_1$ and $z_1$.
Recall the definition of the cross product:
\[
\vec{r}_1 \cross \vec{r}_2 =  \vector{y_1z_2 - z_1y_2, z_1x_2 - x_1z_2, x_1y_2 - y_1x_2}
\]
Computing the cross product and then differentiating yields:
\begin{align*}
\frac{d}{dt}\left[\vec r_1(t) \cross \vec r_2(t)\right] &= \frac{d}{dt} \left[ \vector{y_1z_2 - z_1y_2, z_1x_2 - x_1z_2, x_1y_2 - y_1x_2}\right]\\
&=  \vector{\frac{d}{dt} \left[y_1z_2 - z_1y_2\right], \frac{d}{dt} \left[z_1x_2 - x_1z_2\right],\frac{d}{dt} \left[x_1y_2 - y_1x_2\right]}\\
&= \vector{y_1z_2' + y_1'z_2- z_1y_2'  - z_1'y_2,  z_1x_2' + z_1'x_2- x_1z_2'  - x_1'z_2, x_1y_2' +x_1'y_2- y_1x_2'  - y_1'x_2      }\\
&=\vector{y_1z_2' - z_1y_2'  ,  z_1x_2' - x_1z_2', x_1y_2' - y_1x_2'}+ \vector{ y_1'z_2  - z_1'y_2, z_1'x_2 - x_1'z_2, x_1'y_2 - y_1'x_2}\\
&= \vector{x_1, y_1, z_1}\cross \vector{x_2', y_2', z_2'} + \vector{x_1', y_1', z_1'}\cross \vector{x_2, y_2, z_2}\\
&= \vec r_1(t) \cross \vec r_2\,'(t) + \vec r_1\,'(t) \cross \vec r_2(t)
\end{align*}
as claimed.


\end{proof}

\begin{problem}(Problem 2)
Let $\vec r(t)$ be a vector-valued function whose components are twice differentiable. Use the product rule for the cross product to show that
\[
\frac{d}{dt}\left[\vec r(t) \cross \vec r\,'(t) \right] = \vec r\,'(t) \cross \vec r\,''(t)
\]
\begin{hint}
The cross product of parallel vectors is the zero vector
\end{hint}
\end{problem}

\begin{problem}(Problem 3)
Let $\vec r(t)$ be a vector-valued function whose components are three times differentiable. Use the product rules for both the dot and cross products to show that
\[
\frac{d}{dt}\left[\vec r(t) \dotp \left(\vec r\,'(t) \cross \vec r\,''(t)\right)\right] = \vec r(t) \dotp \left(\vec r\,'(t) \cross \vec r\,'''(t)\right)
\]
\begin{hint}
The cross product of two vectors is orthogonal to both of them
\end{hint}
\begin{hint}
The cross product of parallel vectors is the zero vector
\end{hint}
\end{problem}

\begin{problem}(Problem 4)
Let $\vec u(t), \vec v(t)$ and $\vec w(t)$ be vector-valued functions. Find an expression for
\[
\frac{d}{dt}\left[ \vec u(t) \dotp \left(\vec v(t) \cross \vec w(t)\right)\right]
\]
Compare this to the derivative of a product of three ordinary functions.
\begin{hint}
For ordinary functions $,f, g$ and $h$
\[
\frac{d}{dx}\left[f(x)g(x)h(x)\right] = f(x)g(x)h'(x) + f(x)g'(x)h(x) + f'(x)g(x)h(x)
\]
\end{hint}
\end{problem}

\section{Chain rule}
Recall the chain rule for the derivative of a composition of two differentiable functions $f$ and $g$:
\[
\frac{d}{dx} f(g(x)) = f'(g(x)) \cdot g'(x)
\]

\begin{proposition}[Chain Rule for Vector-Valued Functions]
Let $\vec r(t) = \vector{x(t), y(t),z(t)}$ be a vector-valued function with differentiable components, and let $g(t)$ be an ordinary, differentiable real-valued function.
Then
\[
\frac{d}{dt}\vec r(g(t)) = g'(t)\,\vec r\,'(g(t))
\]
\end{proposition}
\begin{proof}
We write the composition in component form and then use the chain rule on each component:
\begin{align*}
\frac{d}{dt} \vec r(g(t)) &= \frac{d}{dt} \vector{x(g(t)), y(g(t)),z(g(t))}\\
&= \vector{\frac{d}{dt} x(g(t)), \frac{d}{dt} y(g(t)), \frac{d}{dt} z(g(t))}\\
&= \vector{x'(g(t)) g'(t), y'(g(t)) g'(t), z'(g(t)) g'(t)}\\
&= g'(t)\vector{x'(g(t)), y'(g(t)), z'(g(t))}\\
&= g'(t) \vec r\,'(g(t))
\end{align*}
as claimed.
\end{proof}

\begin{problem}(Problem 5)
Let $g(t) = \sqrt t$ and let $\vec r(t) = \vector{t^2, \sec(t), \tan(t)}$. Find
\[
\frac{d}{dt} \vec r(g(t))
\]
\end{problem}
 

\end{document}
