\documentclass[handout]{ximera}

\input{../preamble}

\outcome{In this section we define vectors in two dimensions and study their algebraic and geometric properties.}

\title{1.2 Vectors in $\R^2$}
%Vectors are represented graphically by arrows.
%and in three dimensions we write $\avec{v} = \vector{x,y, z}$.
%The length of the arrow represents the magnitude of the vector and the arrow points in the direction of the vector.
\begin{document}

\begin{abstract}
In this section we define vectors in two dimensions and study their algebraic and geometric properties.
\end{abstract}
 
\maketitle

A {\bf vector} is a quantity that has both magnitude and direction. Examples include a force acting on an object and the velocity of an object in motion.
Algebraically, vectors are represented in a manner very similar to points.  In two dimensions, a vector is written $\avec{v} = \vector{x,y}$  
where $x$ and $y$ are the {\bf components} of $\avec{v}$.


\section{Graphing Vectors in $\R^2$}
The graphical representation of the vector $\avec{v} = \vector{x,y}$ in standard position
is an arrow emanating from the origin (initial point) and terminating at the point $(x,y)$ (final point).
Below is a graphical representation of the vector $\vector{1,1}$ in standard position.

\begin{image}
\begin{tikzpicture}
\draw[ ->] (0,0) -- (3.6, 0) node[right]{$x$};
\draw[ ->] (0,0) -- (0, 3.6) node[above]{$y$};
\draw[->, blue] (0,0) -- (2, 2) node[right]{$\vector{1,1}$};
\draw[blue, fill] (0,0) circle (0.03);
\draw[thin] (2, 0.15) -- (2, -0.15) node[below]{$1$};
\draw[thin] (0.15,2) -- (-0.15,2) node[left]{$1$};
\end{tikzpicture}
\end{image}

It is not necessary to place a vector in standard position.  
In practical applications of vectors, it is typical to place the initial point of a vector in a position of interest.  
For example, if the vector represents the velocity of an object, then one would likely choose the initial point to be the location of the object.
Below is a graphical representation of the vector, $\vector{1,1}$ placed in several different positions.
Since each of the arrows below are the same length and point in the same direction, they represent the same vector.
Notice that in each copy of the vector, the final point is one unit to the right and one unit up from the initial point. 
\begin{image}
\begin{tikzpicture}
\draw[step=1.0,black,thin] (-2.5,-2.5) grid (2.5,2.5);
\draw[thick, <->] (-2.5,0) -- (2.5, 0) node[right]{$x$};
\draw[thick, <->] (0,-2.5) -- (0, 2.5) node[above]{$y$};
\draw[->, blue] (0,0) -- (1, 1) ;
\draw[->, blue] (-2,0) -- (-1, 1) ;
\draw[->, blue] (0,1) -- (1, 2) ;
\draw[->, blue] (-2,1) -- (-1, 2) ;
\draw[->, blue] (1,-2) -- (2, -1) ;
\draw[->, blue] (0,-1) -- (1, 0) ;
\draw[->, blue] (-2,-2) -- (-1, -1) ;
\draw[->, blue] (1,-1) -- (2, 0) ;
\node at (2.2, -0.3){$2$};
\node at (-1.8, -0.3){$-2$};
\node at (-0.3,2){$2$};
\node at (-0.3, -2){$-2$};
\node at (0, -2.8) {Multiple copies of the vector $\vector{1,1}$};
\end{tikzpicture}
\end{image}

Two special vectors in $\R^2$ are $\avec{i} = \vector{1,0}$ and $\avec{j} = \vector{0,1}$.
These vectors both have a magnitude of $1$, making them {\bf unit vectors}. 
Vertical bars are used to indicate magnitude, so $|\avec{i}| = 1$ and $|\avec{j}| = 1$.
The vectors $\avec{i}$ and $\avec{j}$ are shown below.

\begin{image}
\begin{tikzpicture}
\draw[ <->] (-1.5,0) -- (1.8, 0) node[right]{$x$};
\draw[<->] (0,-1.5) -- (0,1.8) node[above]{$y$};
\draw[thick,->, blue] (0,0) -- (1, 0) node[above ]{$\avec{i}$};
\draw[thick,->, blue] (0,0) -- (0,1) node[right]{$\avec{j}$};
\draw[blue, fill] (0,0) circle (0.01);
\draw[thin] (1, 0.1) -- (1, -0.1) node[below]{$1$};
\draw[thin] (0.1,1) -- (-0.1,1) node[left]{$1$};
\draw[white] (-1, -2) -- (1, -2);
\end{tikzpicture}
\end{image}

Another special vector is the zero vector, $\avec{0} = \vector{0,0}$.  
This vector has magnitude zero and no direction.  
It is represented graphically by a single point.

\begin{image}
\begin{tikzpicture}
\draw[ <->] (-2,0) -- (2, 0) node[right]{$x$};
\draw[<->] (0,-2) -- (0, 2) node[above]{$y$};
\draw[blue, fill] (0,0) circle (0.03) node[below right]{$\avec{0}$};
%\draw[thin] (1, 0.15) -- (1, -0.15) node[below]{$1$};
%\draw[thin] (0.15,1) -- (-0.15,1) node[left]{$1$};
\end{tikzpicture}
\end{image}

The vector with initial point $P(x_1, y_1)$ and final point $Q(x_2, y_2)$ has components $\avec{PQ} = \vector{x_2 - x_1, y_2-y_1}$


\begin{image}
\begin{tikzpicture}
\draw[ <->] (-2,0) -- (2, 0) node[right]{$x$};
\draw[ <->] (0,-1) node[below]{The vector $\avec{PQ} = \vector{x_2 - x_1, y_2-y_1}$} -- (0, 2) node[above]{$y$};
\draw[->, blue] (-1,0.5) node[left]{$P(x_1,y_1)$} -- (1.5, 1.5) node[right]{$Q(x_2,y_2)$};
\draw[blue, fill] (-1, 0.5) circle (0.03);
%\draw[thin] (2, 0.15) -- (2, -0.15) node[below]{$1$};
%\draw[thin] (0.15,2) -- (-0.15,2) node[left]{$1$};
\end{tikzpicture}
\end{image}

\begin{example}
Find the components of the vector from the point $P(-2, 3)$ to the point $Q(1, 1)$ and draw two copies of this vector- the first from the point $P$ to the point $Q$
and the second in standard position.\\
The components of $\avec{PQ}$ are $\vector{1-(-2), 1-3} = \vector{3, -2}$.
The two copies of the vector are shown below.

\begin{image}
\begin{tikzpicture}
\draw[step=1.0,black,thin] (-3.5,-3.5) grid (3.5,3.5);
\draw[thick, <->] (-3.5,0) -- (3.5, 0) node[right]{$x$};
\draw[thick, <->] (0,-3.5) -- (0, 3.5) node[above]{$y$};
\draw[->, blue, thick] (0,0) -- (3, -2) ;
\draw[blue, fill] (0, 0) circle (0.03);
\draw[->, blue, thick] (-2,3) node[above left]{$P$} -- (1, 1) node[right]{$Q$} ;
\draw[blue, fill] (-2, 3) circle (0.03);
\node at (2.1, -0.3){$2$};
\node at (-2, -0.3){$-2$};
\node at (-0.3,2){$2$};
\node at (-0.3, -2){$-2$};
\node at (0, -3.8) {Two copies of the vector $\avec{PQ} = \vector{3,-2}$};
\end{tikzpicture}
\end{image}

\end{example}

\begin{problem}
Find the components of the vector from $P(-3, 2)$ to the point $Q(2, -1)$ and draw two copies of this vector- the first from the point $P$ to the point $Q$
and the second in standard position.\\
\end{problem}


\subsection{Magnitude of a Vector}

The magnitude of a vector $\avec{v} = \vector{x,y}$ in $\R^2$ is given by the distance formula:
\[
|\avec{v}| = |\vector{x,y}| = \sqrt{x^2 +y^2}
\]

\begin{example}
Find the magnitude of the vector $\vector{1,1}$.\\
Its magnitude is
\[
|\vector{1,1}| = \sqrt{1^2 +1^2} = \sqrt 2
\]
\end{example}

\begin{problem}
Find the magnitudes of the following vectors:\\
a) $|\vector{2,3}| = \answer{\sqrt{13}}$\\
b) $|\vector{-4,2}| = \answer{2\sqrt{5}}$\\
c) $|\vector{12,-5}| = \answer{13}$
\end{problem}

The magnitude of the zero vector is $0$ and it is the only vector with magnitude zero
\[
|\avec{0}| = |\vector{0,0}| = \sqrt{0^2 +0^2} = 0
\]

\section{Vector Operations}
In this section, we learn about two fundamental operations involving vectors: scalar multiplication and addition.
The more advanced dot product and cross product operations will be handled in future sections.

\subsection{Scalar Multiplication}
Given a vector $\avec{v} = \vector{x,y}$ in $\R^2$ and a scalar $c$ in $\R$, we can multiply the scalar and the vector in the following
natural way
\[
c\avec{v} = c\vector{x,y} = \vector{cx,cy}
\]
In words, we multiply a vector by a scalar by multiplying each component by the scalar.
The first thing to note is that if the scalar is 0, then the resulting vector will be the zero vector:
\[
0\avec{v} = 0\vector{x,y} = \vector{0,0} = \avec{0}
\]

The effect of multiplying a vector by a scalar on its magnitude is given in the following proposition.
\begin{proposition}
Given a vector $\avec{v} = \vector{x,y}$ in $\R^2$ and a scalar $c$ in $\R$,
\[
|c\avec{v}| = |c| |\avec{v}|
\]
\begin{proof}
Write $\avec{v}$ in component form as $\avec{v} = \vector{x,y}$, then
\begin{align*}
|c\avec{v}| &= |c\vector{x,y}|\\
            &= |\vector{cx,cy}|\\
            &= \sqrt{(cx)^2 + (cy)^2}\\
            &= \sqrt{c^2x^2 + c^2y^2}\\
            &= \sqrt{c^2(x^2 + y^2)}\\
            &= \sqrt{c^2} \sqrt{x^2 + y^2}\\
            &= |c| |\avec{v}|
\end{align*}
\end{proof}
\end{proposition}

When $c>0$, the vector $c\avec{v}$ has the same direction as $\avec{v}$, and when $c<0$, the
the vector $c\avec{v}$ has the opposite direction as $\avec{v}$ as illustrated below.



\begin{image}
\begin{tikzpicture}
\draw[blue, thick, ->] (-3,0) -- (-2,0) node[right]{$\avec{v}$};
\draw[red, thick, <-] (-3,-1) node[left ]{$-\avec{v}$} -- (-2,-1) ;
\draw[blue, thick, ->] (0,0) -- (2,0) node[right]{$2\avec{v}$};
\draw[red, thick, <-] (0,-1) node[left]{$-2\avec{v}$} -- (2,-1);
\draw[blue, thick, ->] (4,0) -- (4.5,0) node[right]{$\tfrac12\avec{v}$};
\draw[red, thick, <-] (4,-1)node[left]{$-\tfrac12\avec{v}$} -- (4.5,-1) ;
\node at (0.75, -2) {Multiplying $\avec{v}$ by positive and negative scalars};
%\draw[white] (0,-4) -- (6, -4);
\end{tikzpicture}
\end{image}

A vector with magnitude 1 is called a {\bf unit vector}. Given a vector $\avec{v}$, we will want to find a unit vector in the direction of $\avec{v}$.
This unit vector will have the form $c\avec{v}$ for some positive scalar $c$.  
\begin{proposition}
The unit vector $\avec{u}$ in the same direction as the vector $\avec{v} \neq \avec{0}$ is given by 
\[
\avec{u} = \frac{1}{|\avec{v}|} \avec{v}
\]
\begin{proof}
Since $|\avec{v}| \neq 0$, the scalar $\displaystyle \frac{1}{|\avec{v}|}$ is a positive real number.  Thus, the vector $\avec{u}$ points 
in the same direction as $\avec{v}$, since it is a positive multiple of $\avec{v}$ . We must now show that $|\avec{u}| = 1$.  
By the previous proposition, we have
\begin{align*}
|\avec{u}| &= \left| \frac{1}{|\avec{v}|} \avec{v} \right|\\
           &= \left| \frac{1}{|\avec{v}|} \right| |\avec{v}|\\
           &= \frac{1}{|\avec{v}|} |\avec{v}|\\
           &= 1
\end{align*}
\end{proof}
\end{proposition}

\begin{example}
Find a unit vector in the same direction as the vector $\avec{v} = \vector{3, -4}$.\\
First we find the magnitude of $\avec{v}$:
\[
|\avec{v}| = |\vector{3, -4}| = \sqrt{3^2 + (-4)^2} = \sqrt{25} = 5
\]
By the proposition above, we have
\[
\avec{u} = \frac{1}{|\avec{v}|} \avec{v} = \frac15 \vector{3, -4} = \vector{\frac35, -\frac45}
\]
where $\avec{u}$ is a unit vector in the same direction as $\avec{v}$.
\end{example}

\begin{problem}
Find a unit vector in the same direction as the vector $\avec{v} = \vector{-8, 15}$.\\
$\avec{u} = \vector{\answer{-8/17}, \answer{15/17}}$
\end{problem}

\begin{problem}
Find a unit vector in the opposite direction of the vector $\avec{v} = \vector{9, -12}$.\\
$\avec{u} = \vector{\answer{-3/5}, \answer{4/5}}$
\end{problem}

Every non-zero vector in $\R^2$ can be written in terms of its magnitude and a direction angle, i.e. a polar form.

\begin{proposition}
Let $\avec{v} = \vector{x,y} \neq \vector{0,0}$.  Then there is an angle $\theta$ such that $\avec{v} = |\avec{v}| \vector{\cos \theta, \sin \theta}$.
\begin{proof}
Since $\cos^2\theta + \sin^2\theta = 1$ for any angle $\theta$, the vector $\vector{\cos \theta, \sin \theta}$ is a unit vector.
By the previous proposition, the vector $|\avec{v}| \vector{\cos \theta, \sin \theta}$ has magnitude $|\avec{v}|$. 
Now, if $\avec{v}$ is draw in standard position, and if $\theta$ is the (counterclockwise) angle between $\avec{v}$ and the positive $x$-axis, 
then the components of $\avec{v}$ are $|\avec{v}|\cos \theta$ and $|\avec{v}|\sin \theta$.  See the figure below.


\begin{image}
\begin{tikzpicture}
\draw[step=1.0,black,thin] (-3.5,-3.5) grid (3.5,3.5);
\draw[thick, <->] (-3.5,0) -- (3.5, 0) node[right]{$x$};
\draw[thick, <->] (0,-3.5) -- (0, 3.5) node[above]{$y$};
\draw[->, blue, thick] (0,0) -- (-2.5, 2.5) node[above, right]{$\avec{v} = \vector{|\avec{v}|\cos\theta,|\avec{v}|\sin\theta}$};
\draw[blue, fill] (0, 0) circle (0.03);
\draw[red] (0,0) circle (1);
\draw[red, fill] (-.707, .707) circle (0.03) node[above, left]{$(\cos \theta, \sin\theta)$};
\draw[red] (.2,0) arc (0:135:0.2) node[red, midway, right]{$\theta$};
\node at (2.1, -0.3){$2$};
\node at (-2, -0.3){$-2$};
\node at (-0.3,2){$2$};
\node at (-0.3, -2){$-2$};
\node[red] at (1.3, -1.5) {The Unit Circle};
%\node at (0, -3.8) {Two copies of the vector $\avec{PQ} = \vector{3,-2}$};
\end{tikzpicture}
\end{image}


\end{proof}
\end{proposition}

\begin{example}
Find the polar form of the vectors $-2\avec{i}$ and $4\avec{i}-7\avec{j}$.\\
The angle associate with the vector $-2\avec{i}$ is $\pi$ since the vector points in the direction of the negative $x$-axis.
So $-2\avec{i} = |-2\avec{i}| \vector{\cos \pi, \sin \pi} = 2\vector{\cos \pi, \sin \pi} (= 2\vector{-1, 0} = \vector{-2, 0})$.\\
The angle associated with the vector $4\avec{i}-7\avec{j}$ is $\theta = \tan^{-1}(-7/4)$, which is not a well known angle.
The magnitude of $4\avec{i}-7\avec{j}$ is $\sqrt{16 +49} = \sqrt{65}$.  Hence, the polar form of $4\avec{i}-7\avec{j}$ is
$\sqrt{65}\vector{\cos \theta, \sin\theta}$ where $\theta$ is as above. Note that if $\avec{v}$ is in the second or third quadrant, then 
to find the angle we would add $\pi$ to the inverse tangent.
\end{example}

\begin{problem}
Find the polar form of the vectors $-3\avec{j}$ and $-2\avec{i}+5\avec{j}$.\\
\end{problem}

\subsection{Vector Addition}
Vector addition is done component-wise.  If $\avec{v_1} = \vector{x_1, y_1}$ and $\avec{v_2} = \vector{x_2, y_2}$
then their sum is given by
\[
\avec{v_1 }+ \avec{ v_2}= \vector{x_1+x_2,  y_1+y_2}
\]
Vector subtraction is addition of the negative which results in component-wise subtraction:
\[
\avec{v_1 }- \avec{ v_2}= \avec{v_1 }+ \left(- \avec{ v_2}\right)  = \vector{x_1,y_1} + \vector{-x_2, -y_2} = \vector{x_1-x_2,  y_1-y_2}
\]

\begin{example}
Let $\avec{u} = \vector{-4, 2}$ and $\avec{v} = \vector{5, -7}$.  Compute each of the following:\\
a) $\avec{u}+\avec{v} = \vector{1, -5}$\\
b) $\avec{u}-\avec{v} = \vector{-9, 9}$\\
c) $\avec{v}-\avec{u} = \vector{9, -9}$\\
d) $2\avec{u}+3\avec{v} = \vector{7, -17}$\\
\end{example}

\begin{problem}
Let $\avec{u} = \vector{2, -3}$ and $\avec{v} = \vector{-1, 6}$.  Compute each of the following:\\
a) $\avec{u}+\avec{v} = \vector{\answer{1}, \answer{3}}$\\
b) $\avec{u}-\avec{v} = \vector{\answer{3}, \answer{-9}}$\\
c) $\avec{v}-\avec{u} = \vector{\answer{-3}, \answer{9}}$\\
d) $3\avec{u}+5\avec{v} = \vector{\answer{1}, \answer{21}}$\\
\end{problem}

Graphically, vector addition can be performed using either the parallelogram method or the end to end method as pictured below.
\begin{image}
\begin{tikzpicture}
\draw[blue, ->, thick] (0,0) -- (3, 1) node[midway, below]{$\avec{u}$};
\draw[blue, ->, thick] (0,0) -- (1, 4) node[midway, left]{$\avec{v}$};
\draw[blue, thin, dashed] (3, 1) -- (4, 5);
\draw[blue, thin, dashed] (1, 4) -- (4, 5);
\draw[red, ->, thick] (0,0) -- (4, 5) node[above]{$\avec{u} + \avec{v}$};
\node at (1.5, -1){Parallelogram Method};
\draw[blue, ->, thick] (6,0) -- (9, 1) node[midway, below]{$\avec{u}$};
\draw[blue, ->, thick] (9,1) -- (10, 5) node[midway, right]{$\avec{v}$};
\draw[red, ->, thick] (6,0) -- (10, 5) node[above]{$\avec{u} + \avec{v}$};
\node at (7.5, -1){End to End Method};
\end{tikzpicture}
\end{image}

The difference $\avec{u} - \avec{v}$ is the vector from $\avec{v}$ to $\avec{u}$ as shown below.
\begin{image}
\begin{tikzpicture}
\draw[blue, ->, thick] (0,0) -- (3, 1) node[midway, below]{$\avec{u}$};
\draw[blue, ->, thick] (0,0) -- (1, 4) node[midway, left]{$\avec{v}$};
\draw[red, ->, thick] (1,4) -- (3, 1) node[midway,right]{$\avec{u} - \avec{v}$};
\node at (1.5, -1){The vector $\avec{u} - \avec{v}$};
\draw[white] (-3, -1.5) -- (7, -1.5);
\end{tikzpicture}
\end{image}
Note that in the figure above, if we add the vectors $\avec{v}$ and $\avec{u} -\avec{v}$ using the end to end method, 
we get the vector $\avec{u}$, i.e.,
\[
\avec{v} + \left(\avec{u} -\avec{v}\right) = \avec{u}
\]

Vectors in $\R^2$ can be written using the {\bf standard basis} vectors $\avec{i} = \vector{1,0}$ and $\avec{j} = \vector{0,1}$.  
Given a vector $\avec{v} = \vector{x,y}$, we can use the addition and scalar multiplication operations to rewrite it as
\begin{align*}
\avec{v} &= \vector{x,y}\\
         &= \vector{x,0} + \vector{0,y}\\
         &= x\vector{1,0} + y \vector{0,1}\\
         &= x\avec{i} + y\avec{j}
\end{align*}



\section{Application}
Forces acting on an object have both magnitude and a direction and hence they can be modeled using vectors.

\begin{example}
A box on the floor is being pulled to the right by a rope. The weight of the box is $20$ Newtons, the force due to friction is $8$ Newtons, and the 
force on the rope is $24$ Newtons at an angle of $30^\circ$ to the floor. Find the normal force, $\avec{N}$, acting on the box from the 
floor and resultant force acting on the box, $\avec{F}$.\\
\begin{image}  
\begin{tikzpicture}
\draw (0,0) -- (6,0) node[right]{floor};
\draw (1.5, 0) -- (1.5, 2) -- (4.5, 2) -- (4.5, 0) node[midway, right]{box};
\draw[red, ->] (3,0) -- (3, -1) node[below]{weight = $20$N};
\draw[red, <-] (1, 1) node[left]{friction = $8$N} -- (1.5,1);
\draw[red, ->] (4.5, 2) -- (6, 3) node[right]{rope force = $24$N};
\draw[red, ->] (3,2) -- (3, 2.5) node[above]{normal force, $\avec{N}$};
\draw[red, dashed] (4.5, 2) -- (5.7, 2) ;
\node[red] at (5.3,2.2){$30^\circ$};
\end{tikzpicture}
\end{image}


The resultant force on the box is the sum of the forces acting on the box, where each force is represented by a vector.
The force of the weight of the box has a magnitude of $20$ and points in the direction of the negative $y$-axis.  
This vector is $\avec{w} = \vector{0,-20}$.
The friction force has a magnitude of $8$ and points in the direction of the negative $x$-axis.  This vector is $\avec{f} = \vector{-8,0}$.
The force on the rope has a magnitude of $24$ and points in the direction $30^\circ$ above the $x$-axis. 
Using trigonometry, we can deduce that the horizontal and vertical components of this force vector are $x = 24\cos(30^\circ) = 12\sqrt 3$
and $y = 24\sin(30^\circ) = 12$ respectively. Hence, this force is $\avec r = \vector{12\sqrt 3, 12}$.  See the figure below.

\begin{image}
\begin{tikzpicture}
\draw(0,0) -- (6.8,0) node[midway, below]{$24 \cos(30^\circ)$} ;
\draw (6.8, 0) -- (6.8,4) node[midway,right]{$24\sin(30^\circ)$};
\draw[red, ->] (0,0) -- (6.8, 4) node[midway, above left]{$|\avec{r}| = 24$};
\node at (1, 0.25) {$30^\circ$};
\node at (3.4, -1) {The force on the rope is $\avec r = \vector{12\sqrt 3, 12}$};
\draw (6.5, 0) -- (6.5, 0.3) -- (6.8, 0.3);
\end{tikzpicture}
\end{image}

Assuming that the vertical component of the rope force is not enough to overcome the weight of the box (and thus send it airborne), 
the floor will exert an upward normal force on the box.  This force will have the form $\avec{N} = \vector{0, n}$ where the positive 
number $n$ will be determined by the fact that the net vertical component of all forces acting on the box is zero. 
The resultant force, $\avec{F}$, on the box is the sum of the vectors $\avec{f}, \avec{w}, \avec{N}$ and $\avec{r}$:
\begin{align*}
\avec{F} &= \avec{f} + \avec{w} + \avec{N} +  \avec{r} \\
         &= \vector{-8,0} + \vector{0,-20} +\vector{0,n} + \vector{12\sqrt 3, 12} \\
         &= \vector{12\sqrt 3 -8 , n-8}
\end{align*}
Since the vertical component of the resultant force $\avec{F}$ should be zero, we see that $n = 8$. 
Thus $\avec{N} = \vector{0,8} = 8\avec{j}$ so that $\avec{F} = \vector{12\sqrt 3 -8, 0} \approx \vector{12.785,0} = 12.785\avec{i}$. Because of this force, 
the box will slide along the floor to the right.

% at a speed equal to the magnitude of this force: 
%\[
%|\avec{F}| = \left|\vector{12\sqrt 3 -8, 0}\right| = 12\sqrt 3 - 8 \approx 12.785\, \text{meters/second}
%\]




\end{example}
\end{document}

\begin{image}  
\begin{tikzpicture}

\draw[red, ->] (0,0) -- (0, -1) node[below]{W\\avec{w} = \vector{0, -20}};
\draw[red, <-] (1, 1) node[left]{friction = $8$N} -- (1.5,1);
\draw[red, ->] (4.5, 2) -- (6, 3) node[right]{rope force = $25$N};
\draw[red, dashed] (4.5, 2) -- (5.7, 2) ;
\node[red] at (5.3,2.2){$30^\circ$};
\end{tikzpicture}
\end{image}










\begin{image}
\begin{tikzpicture}
\draw[blue, thick, ->] (-3,0) -- (-2,0) node[right]{$\avec{v}$};
\draw[blue, thick, <-] (-3,-0.5) node[left ]{$-\avec{v}$} -- (-2,-0.5) ;
\draw[red, thick, ->] (0,-1) -- (2,-1) node[right]{$2\avec{v}$};
\draw[red, thick, <-] (0,-1.5) node[left]{$-2\avec{v}$} -- (2,-1.5);
\draw[green, thick, ->] (0,-2) -- (0.5,-2) node[right]{$\tfrac12\avec{v}$};
\draw[green, thick, <-] (0,-2.5)node[left]{$-\tfrac12\avec{v}$} -- (0.5,-2.5) ;

\end{tikzpicture}
\end{image}



\begin{image}
\begin{tikzpicture}
\draw[thick, ->] (0,0) -- (2,0) node[right]{$y$};
\draw[thick, ->] (0,0) -- (0,1.8) node[above]{$z$};
\draw[thick, ->] (0,0) -- (-1.2,-0.9) node[below, left]{$x$};
\draw[blue] (0.3,0.2) -- (1.7, 0.5) -- (1.7, 1.7) -- cycle;
\draw[thin, blue]  (1.6, 0.47) -- (1.6, .59) -- (1.7, .62);
\draw[fill, blue] (0.3,0.2) circle [radius=0.03] node[left, blue]{$(x_1, y_1, z_1)$};
\draw[fill, blue] (1.7,1.7) circle [radius=0.03] node[right, blue]{$(x_2, y_2, z_2)$};
\draw[fill, blue] (1.7,0.5) circle [radius=0.037] node[right, blue]{$(x_2, y_2, z_1)$};
\node[below, blue] at (1.3,0.4) {$d_1$}; %adjacent
\node[right, blue] at (1.9,1) {$d_2$}; %opposite 
%\node[right] at (4.1,2) {$\tan(\theta)$};
\node[above, blue] at (1,1.4) {$d$}; %hypotenuse
%\node[below] at (1.75,-0.6) {$d = \sqrt{(x_2 - x_1)^2 + |y_2 - y_1|^2}$};
%\node[above] at (1.75,3.6) {A right triangle with $\cos(\theta) = x$};
\end{tikzpicture}
\end{image}
