\documentclass{ximera}

\input{../preamble}

\outcome{Find the volume of a solid}

\title{Volumes of solids}

\begin{document}

\begin{abstract}
We find the volumes of a variety of solids.
\end{abstract}

\maketitle

\begin{center}
\textbf{Volume of Solids}
\end{center}

We have learned that the integral can be used to compute areas of regions in the plane.  
We will now learn that integrals can be used to compute volumes as well.
The integral notation suggests that we compute area in the plane by summing the areas of  infinitely thin line segments.
Of course this perspective is only the result of a limit of Riemann Sums of areas of rectangles.
The volume of a solid can be similarly conceptualized as resulting from pasting together infinitely many cross-sections,
each of which possesses area. Hence the integral can be used in a similar fashion to compute the volume of a solid.
In the case of volume, the defining Riemann Sums involve irregularly formed cylinders rather than rectangles
in our approximation process.

%replacing the line segments which when pasted together form a region,
%with cross-sectional areas which when themselves pasted together form a solid.

%visual aid here

Suppose that a solid be placed atop an axis, say the $x-$axis, as shown in the figure below.

%visual aid here

For each $x$ between $a$ and $b$, there is a cross-section of the solid whose area we will denote by $A(x)$.
Then our astonishing result is that the volume of the solid can be computed as

\[
V = \int_a^b A(x) \; dx.
\]

\begin{remark}
We can also re-orient our axis so that it would be more aptly, named $y$, and with respect to this vertical axis, 
we could compute the volume with the exact same idea as
\[
V = \int_a^b A(y) \; dy.
\]

\end{remark}

%visual aid here with y-axis


\begin{explanation}
To be precise, we will next define the volume in terms of the limit of a Riemann Sums.  
Let us reconsider the solid above, from the perspective of the x-axis. Suppose that at
each $x-value$ there is a cross-sectional area $A(x)$. Then to approximate the volume of the solid we 
can use a Riemann sum as follows. Subdivide $[0,1]$ into $n$ equal sub-intervals each of length $\Delta x = 1/n$.
For each sub-interval (whose form is $[\frac{i-1}{n}, \frac{i}{n}]$) we choose a sample point and call it $x_i^*$, where 
the index $i$ runs from $1$ to $n$. 
Corresponding to these $n$ values of $x$, there are cross-sections whose areas are given by $A(x_i^*)$.
From these cross-sections we can construct an irregular cylinder as shown below.

%visual aid here of irregular cylinder

The volume of each of these $n$ cylinders is given by
\[
V = \text{area of base} \times \text{height} = A(x_i^*) \cdot \frac{1}{n} = A(x_i^*) \cdot \Delta x.
\]
Summing the volumes of these $n$ cylinders gives an approximation to the volume as a Riemann Sum
\[
V \approx \sum_{i=1}^n A(x_i^*) \Delta x.
\]
The exact volume is then defined in terms of a limit these Riemann Sums as $n$ tends to infinity,
\[
V = \int_0^1 A(x) \; dx = \lim_{n \to \infty} \sum_{i=1}^n A(x_i^*) \Delta x.
\]

\end{explanation}

We generalize this explanation in the definition below.

\begin{definition}[Volume of a Solid]
The exact volume of a solid whose cross-sectional area at an $x$ value between $a$ and $b$ is given by
\[
V = \int_a^b A(x) \; dx = \lim_{n \to \infty} \sum_{i=1}^n A(x_i^*) \Delta x.
\]
In this context, 
\[
\Delta x = \frac{b-a}{n}
\] 
and $x_i^*$ is a sample point in the $i$-th sub-interval
\[
[a + (i-1)\Delta x, a + i\Delta x],
\]
where $i = 1, 2, \dots, n$.
\end{definition}















































\begin{center}
\begin{foldable}
\unfoldable{Here is a video of Example 1}
%\youtube{} %vid of example 1
\end{foldable}
\end{center}




\end{document}




%multiple choice format; note the question environment; what is free response? what is verbatim?

%\begin{verbatim}
\begin{question}
What is your favorite color?
\begin{multipleChoice}
\choice[correct]{Rainbow}
\choice{Blue}
\choice{Green}
\choice{Red}
\end{multipleChoice}
\begin{freeResponse}
Hello
\end{freeResponse}
\end{question}
%\end{verbatim}


%note the indentation

\begin{question}
  Which one will you choose?
  \begin{multipleChoice}
    \choice[correct]{I'm correct.}
    \choice{I'm wrong.}
    \choice{I'm wrong too.}
  \end{multipleChoice}
\end{question}


%selectAll

\begin{question}
  Which one will you choose?
  \begin{selectAll}
    \choice[correct]{I'm correct.}
    \choice{I'm wrong.}
    \choice[correct]{I'm also correct.}
    \choice{I'm wrong too.}
  \end{selectAll}
\end{question}


\begin{freeResponse}
What is the chain rule used for?
\end{freeResponse}
