\documentclass[handout]{ximera}

\input{../preamble}

\pgfplotsset{compat=1.13}

\outcome{Introduce complex logarithms}

\title{2.2 Complex Logarithms}

\begin{document}

\begin{abstract}
We define and discuss complex logarithms.
\end{abstract}

\maketitle

\section{The Multivalued Logarithm}
Consider the function $w = f(z) = e^z = e^x \cis y$. This function is $2\pi i$-periodic, so it is not one-to-one.  Hence it does not have a traditional inverse-
the complex logarithm is multivalued.  
%For a given $w \in \C$, $z=\log w$ gives all of the values of $z$ so that $e^{z} = w$. 
To define the complex log, consider a complex number $w_0$ in the image of $f(z) = e^z$, 
so that $w_0 = e^{x_0} \cis (y_0+ 2k\pi)$.  Then the modulus of $w_0$ is $e^{x_0}$ and an argument of $w_0$ is $y_0$.


\begin{image}
\begin{tikzpicture}
\draw[<->, thick] (-3,0)--(1,0);

\draw[<->, thick] (-1, 2)--(-1,-2) ;

\draw[mark=*,mark size=1pt,mark options={color=blue}] plot coordinates {(-0.5,.4)} node[right, blue]{$z_0 = x_0 + iy_0$};
\draw[mark=*,mark size=1pt,mark options={color=blue}] plot coordinates {(-0.5,1.4)} node[right, blue]{$z_0+ 2\pi i$};


\draw[mark=*,mark size=1pt,mark options={color=blue}] plot coordinates {(-0.5,-.6)} node[right, blue]{$z_0- 2\pi i$};
\draw[mark=*,mark size=1pt,mark options={color=blue}] plot coordinates {(-0.5,-1.6)} node[right, blue]{$z_0- 4\pi i$};


\draw[<->, thick] (5,0)--(9,0);

\draw[<->, thick] (7, 2)--(7,-2) ;

\draw[mark=*,mark size=1pt,mark options={color=blue}] plot coordinates {(8.5,1.3)} node[right, blue]{$w_0$};
\draw[dashed] (7,0) -- (8.5, 1.3) node[above, midway]{$e^{x_0}$};
\draw (7.4,0) arc (0:42:.4) node[midway, right]{$y_0$} ;
\draw[->, thick, blue] (2.7,0)--(4.3, 0) node[above,midway, blue]{$f(z) = e^z$};
\node at (3, -2.5){\large Note the modulus and argument of $w_0 = e^{z_0 + 2k\pi i}$};
\end{tikzpicture}
\end{image}




We wish to recover $x_0$ and $y_0$.  Thus we define $\log w_0$ so that $\Rep(\log w_0) = \ln|w_0| = \ln e^{x_0} = x_0$ and $\Imp(\log w_0) = \arg w_0 = y_0 + 2k\pi i$. 
In the definition below, we use $z$ in the place of $w_0$.
\begin{definition}[Multivalued Logarithm]
For $z \neq 0$, we define the multivalued complex logarithm by
\[
\log z = \ln |z| + i \arg z
\]
\end{definition}

\begin{example}[example 1]
Find the values of $\log(1+i)$\\
We have 
\[
\log(1+i) = \ln |1+i| + i \arg(1+i) = \ln \sqrt 2 + i\left(\frac{\pi}{4} + 2k\pi\right)
\]
where $k \in \Z$.
\end{example}

\begin{problem}(problem 1a)
Find all values of the complex logarithm: $\; \log i$.\\
\begin{multipleChoice}
\choice{$\frac{\pi}{2} + 2k\pi$}\\
\choice{$1 + i\left(\frac{\pi}{2} + 2k\pi\right)$}\\
\choice[correct]{$i\left(\frac{\pi}{2} + 2k\pi\right)$}\\
\choice{$2k\pi i$}
\end{multipleChoice}
\end{problem}


\begin{problem}(problem 1b)
Find all values of the complex logarithm: $\; \log(-3)$\\
\begin{multipleChoice}
\choice{$ (2k+1)\pi + i\ln 3$}\\
\choice[correct]{$\ln 3 + (2k+1)\pi i$}\\
\choice{$3 + (2k+1)\pi i$}\\
\choice{$(2k-1)\pi i$}
\end{multipleChoice}
\end{problem}


\begin{problem}(problem 1c)
Find all values of the complex logarithm: $\; \log(i-1)$\\
\begin{multipleChoice}
\choice{$\left(\frac{3\pi}{4} + 2k \pi \right) + i \ln \sqrt 2$}\\
\choice{$ \ln \sqrt 2 + i\left(\frac{3\pi}{4} + k \pi \right)$}\\
\choice{$\ln 2 + i\left(\frac{3\pi}{4} + 2k \pi \right)$}\\
\choice[correct]{$\tfrac12 \ln 2 + i\left(\frac{3\pi}{4} + 2k \pi \right)$}
\end{multipleChoice}
\end{problem}



Some familiar properties of real logarithms carry over tho the multivalued complex logarithm.

\begin{proposition}
For $z_1, z_2 \neq 0$, 
\[
\log(z_1\cdot z_2) = \log z_1 + \log z_2
\]
and for $z \neq 0$
\[
\log\left(\frac{1}{z}\right) = - \log z
\]
\end{proposition}
The key to the proof of the first propoerty is that the argument of a product is the sum of the arguments:
\[
\arg(z_1 \cdot z_2) = \arg z_1 + \arg z_2
\]
As a result of this property of the multivalued argument, we have
\begin{align*}
\log(z_1\cdot z_2) &= \ln|z_1 z_2| + i\arg(z_1 z_2)\\
&= \ln|z_1| + \ln|z_2| + i\left[\arg z_1 + \arg z_2\right]\\
&= \log z_1 + \log z_2
\end{align*}

The proof of the second property is similar (verify) and uses
\[
\arg\left(\frac{1}{z}\right) = -\arg z
\]

\begin{corollary}
For $z_1, z_2 \neq 0$
\[
\log\left(\frac{z_1}{z_2}\right) = \log z_1 - \log z_2
\]
\end{corollary}

\section{Principal Log}

Remarking that the complex exponential function, $e^z$, is defined in terms of the radial part, $e^x=r$,
and an angular part $e^{iy}= \cis y$, it is natural that the complex logarithm be defined in terms of the real natural logarithm of a modulus,
and an argument function.

\begin{definition}[$\Log z$]
For $z \neq 0$, we define the principal branch of the complex logarithm by
\[
\Log z =  \ln|z| + i\Arg z
\]
where $\Arg z$ is the principal argument of $z$. 
\end{definition}

 

First, we will show that $\Log z$ and $e^z$ are inverses.


\[
e^{\Log z} = e^{\ln |z| + i\Arg z} = e^{\ln |z|} \cis \Arg z = |z| \cis \Arg z =z
\]
in polar form (if $z=r\cis \theta$ then $r = |z|$ and $\theta \in \arg z$).
Furthermore,
\[
\Log e^z = \Log e^x\cis y = \ln e^x + i\Arg(\cis y) = x+iy
\]
if and only if $-\pi < y \leq \pi$.




\begin{example}[Example 2] 
Find $\Log z$ for $z = -5, i, 1+i, 1-i$ and $-3-i\sqrt 3$.\\
\begin{align*}
& \Log(-5) = \ln|-5|+ i \Arg(-5) = \ln 5 + i\pi\\
& \Log(i) = \ln|i|+ i \Arg(i) = \ln 1 + i\pi/2 = i\pi/2\\
& \Log(1+i) = \ln|1+i|+ i \Arg(1+i) = \ln \sqrt 2 + i\pi/4\\
& \Log(1-i) = \ln|1-i|+ i \Arg(1-i) = \ln \sqrt 2 - i\pi/4\\
& \Log(-3-i\sqrt 3) = \ln|-3-i\sqrt 3|+ i \Arg(-3-i\sqrt 3) = \ln 2\sqrt 3 + i7\pi/6
\end{align*}
\end{example}


\begin{problem}(problem 2) 
Find $\Log z$ for $z = -2, -i, -2-2i, -4+4i$ and $1+i\sqrt 3$.\\
\begin{align*}
& \Log(-2) = \answer{\ln 2 + i\pi }\\
& \Log(-i) = \answer{-i\pi/2}\\
& \Log(-2-2i) = \answer{\ln 2\sqrt 2 - i3\pi/4}\\
& \Log(-4+4i) = \answer{\ln 4\sqrt 2 +i3\pi/4}\\
& \Log(1+i\sqrt 3) = \answer{\ln 2+ i\pi/3}
\end{align*}
\end{problem}

The properties of the multivalued complex logarithm do not carry over to the principal branch without modification.
\begin{proposition}
For $z_1, z_2 \neq 0$
\[\Log(z_1\cdot z_2) = \Log z_1 + \Log z_2
\]
if and only if 
\[
-\pi < \Arg z_1 + \Arg z_2 \leq \pi\]

and for $z \neq 0$
\[
\Log\left(\frac{1}{z}\right) = - \Log z
\]
if and only if $z$ is not a negative real number.
\end{proposition}
\begin{question}Under what conditions can we expect
\[
\Log\left(\frac{z_1}{z_2}\right) = \Log z_1 - \Log z_2?
\]
\end{question}


\section{Other Branches}

Let $\alpha \in \R$ and consider the interval $(\alpha, \alpha + 2\pi]$. For any non-zero $z \in \C$, there is an element of the
set $\arg z$ which belongs to this interval. Hence, we can define a branch of the complex logarithm as follows.

\begin{definition}
Let $\alpha \in \R$ and let $\theta_\alpha$ be an argument of $z$ in the interval $(\alpha, \alpha + 2\pi]$.
The we define
\[
\log_\alpha z = \ln|z| +i\theta_\alpha
\]
\end{definition}


\begin{example}[Example 3] 
Find the indicated logarithm. The expression $\arg_\alpha(z)$ is used below to indicate the element of the set 
$\arg z$ which lies in the interval $(\alpha, \alpha + 2\pi]$.\\
\begin{align*}
i)  \, & \log_0(5) = \ln|5|+ i \arg_0(5) = \ln 5 + i2\pi\\
ii) \,  & \log_\pi(i) = \ln|i|+ i \arg_\pi(i) = \ln 1 + i5\pi/2 = i5\pi/2\\
iii) \,  & \log_{2\pi}(-2) = \ln|-2|+ i \arg_{2\pi}(-2) = \ln 2 + i3\pi\\
iv) \,  & \log_0(-1-i)= \ln|-1-i|+ i \arg_0(-1-i) = \ln \sqrt 2 + i5\pi/4\\
v)  \, & \log_{4\pi}(\sqrt 3 + i) = \ln|\sqrt 3 + i|+ i \arg_{4\pi}(\sqrt 3+i) = \ln 2 + i25\pi/6
\end{align*}
\end{example}


\begin{problem}(problem 3) 
Find the following complex logarithms:\\
\begin{align*}
i) \, & \log_0(3) = \answer{\ln 3 + i2\pi }\\
ii)  \, & \log_\pi(-3) = \answer{\ln 3 + i3\pi}\\
iii) \,  & \log_\pi(-i) = \answer{ i7\pi/2}\\
iv)  \, & \log_{2\pi}(1+i) = \answer{\ln \sqrt 2 +i11\pi/4}\\
v)  \, & \log_0(1- i\sqrt 3) = \answer{\ln 2+ 5\pi/3}
\end{align*}
\end{problem}





\section{Complex Powers}
We define power functions with complex powers. To do so, recall that the real functions $e^x$ and $\ln x$ 
are inverses, so that for $x>0$, we can write
\[
x = e^{\ln x}
\]
Furthermore, a property of logs allows us to write
\[
\ln(x^n) = n \ln x
\]
Combining these gives
\[
x^n = e^{n\ln x}
\]
\begin{definition}
For $c \in \C$ and $z \neq 0$, we define
\[
z^c = e^{c \log z}
\]
and the principal branch of the complex power function is given by
\[
z^c = e^{c \Log z}
\]
\end{definition}

\begin{example}[example 4]
Find all values of $i^i$ and highlight the principal value.\\
Using the definition of a complex exponent, we have
\[
i^i = e^{i\log i} = e^{i(\pi/2 + 2k\pi)i} 
\]
\[
= e^{-(\pi/2 + 2k\pi)}
\]
The principal value occurs when $k=0$.  It is
\[
i^i = e^{-\pi/2} \approx 0.20788...
\]
Note that the values of $i^i$ are real!
\end{example}

\begin{problem}(problem 4)
Find all values of each of the exponential expressions and highlight the principal value.\\
\begin{align*}
i) &\; 4^i; \;\; \mbox{Principal value:}\;\; \answer{e^{i\ln 4}}\\
ii) & \;(1+i)^{\pi i}; \;\;\mbox{Principal value:}\;\; \answer{e^{-\pi^2/4 + \pi i \ln(\sqrt2)}}\\
iii) & \;(-1)^{\sqrt 2}; \;\;\mbox{Principal value:}\;\; \answer{e^{i\pi\sqrt 2}}
\end{align*}
\end{problem}




\end{document}












Let $w = f(z) = z^2$. Find $u(x,y)$ and $v(x,y)$.\\
Since $z = x+iy$, 
\[
z^2 = (x+iy)^2 = \left(x^2 - y^2\right) + i(2xy)
\]
Thus $u(x,y) = x^2 - y^2$ and $v(x,y) = 2xy$.

\end{example}




\begin{problem}(Problem 1a)
Find the real and imaginary parts of the function $w = f(z) = \overline{z}$
\[
u(x,y) = \answer{x} \;\; \mbox{and} \;\; v(x,y) = \answer{-y}
\]

\end{problem}


\begin{problem}(Problem 1b)
Find the real and imaginary parts of the function $w = f(z) = 2iz+i$     
\[
u(x,y) = \answer{-2y} \;\; \mbox{and} \;\; v(x,y) = \answer{2x+1}
\]

\end{problem}


\begin{problem}(Problem 1c)
Find the real and imaginary parts of the function $w = f(z) = z^3$   
\[
u(x,y) = \answer{x^3 - 3xy^2} \;\; \mbox{and} \;\; v(x,y) = \answer{3x^2y -y^3}
\]

\end{problem}

\begin{problem}(Problem 1c)
Find the real and imaginary parts of the function $w = f(z) = 1/z$   
\[
u(x,y) = \answer{x^3 - 3xy^2} \;\; \mbox{and} \;\; v(x,y) = \answer{3x^2y -y^3}
\]

\end{problem}

\section{The Complex Exponential}

To define $e^z$ we will use the power series representations of $e^x, \sin(x)$ and $\cos(x)$:
\[
e^x = \sum_{n=0}^\infty \frac{x^n}{n!},
\]
\[
\sin(x) = \sum_{n=0}^\infty (-1)^n \frac{x^{2n+1}}{(2n+1)!},
\]
and
\[
\cos(x) = \sum_{n=0}^\infty (-1)^n \frac{x^{2n}}{(2n)!}.
\]

We would like the complex exponential function to preserve as many properties of the real exponential function as possible. 
In particular, we would like the complex exponential to have the fundamental property $e^{z_1}\cdot e^{z_2} = e^{z_1 + z_2}$.
In particular, we would like
\[
e^z = e^{x+iy} = e^x \cdot e^{iy}
\]
This reduces the challenge to defining $e^{iy}$. To do this we make a ``formal" power series expansion:
\begin{align*}
e^{iy} &= \sum_{n=0}^\infty \frac{(iy)^n}{n!} = \sum_{n=0}^\infty \frac{i^ny^n}{n!}\\
       &= 1+iy -\frac{y^2}{2!} -i\frac{y^3}{3!} + \frac{y^4}{4!} + i\frac{y^5}{5!} - \frac{y^6}{6!} -i \frac{y^7}{7!} + \cdots\\
       &= \sum_{n=0}^\infty (-1)^n\frac{y^{2n}}{(2n)!} + i\sum_{n=0}^\infty (-1)^n\frac{y^{2n+1}}{(2n+1)!}\\
       &= \cos(y) + i\sin(y)\\
       &= \cis(y)
\end{align*}


In light of this calculation, we define the complex exponential function as
\[
e^z \equiv e^x \cis(y) 
\]
where $z = x+iy$.

\begin{example}[Example 2- Euler's Identity]
Consider the function $f(z) = e^z$. Compute $f(\pi i)$.\\
Writing $\pi i$ as $0+\pi i$, we use the definition of $e^z$ with $x = 0 $ and $y= \pi$.
We have 

\[ f(\pi i) = e^{\pi i} = e^0 \cis \pi = \cos(\pi) + i\sin(\pi) = -1 \]

This gives us \link[Euler's identity]{https://en.wikipedia.org/wiki/Euler's_identity},
\[
e^{\pi} +1 = 0
\]

\end{example}


\begin{problem}(Problem 2)
Calculate each of the following:\\
i) $e^0 = \answer{1}$\\
ii) $e^{\pi i /4} = \answer{1/\sqrt 2 + i/\sqrt 2}$\\
iii) $e^{2\pi i}= \answer{1}$\\
iv) $e^3 = \answer{e^3}$\\
v) $e^{3+2\pi i} = \answer{e^3}$\\
\end{problem} 

\begin{remark}
Since $e^{2\pi i} = 1$, the function $e^z$ is $2\pi i$ periodic.
\end{remark}

%re-write intro sentence to u(x,y) and v(x,y): 
Since $e^z = e^x \cis y$, we have $e^z = u(x,y) + iv(x,y)$ with
\[
u(x,y) = e^x\cos(y) \;\; \mbox{and} \;\; v(x,y)= e^x\sin(y)
\]




\begin{example}[Example 3] 
Let $w = f(z) = e^{2z}$. Find $u(x,y)$ and $v(x,y)$.\\
Since $z = x+iy$, 
\[
e^{2z} = e^{2x+2iy} = e^{2x}\cis(2y)
\]
Thus $u(x,y) = e^{2x}\cos(2y)$ and $v(x,y) = e^{2x}\sin(2y)$.

\end{example}



\begin{problem}(Problem 3a)
Find the real and imaginary parts of the function $w = f(z) = e^{3z}$
\[
u(x,y) = \answer{e^{3x}\cos(3y)} \;\; \mbox{and} \;\; v(x,y) = \answer{e^{3x}\sin(3y)}
\]

\end{problem}

\begin{problem}(Problem 3b)
Find the real and imaginary parts of the function $w = f(z) = e^{-z}$
\[
u(x,y) = \answer{e^{-x}\cos(y)} \;\; \mbox{and} \;\; v(x,y) = \answer{-e^{-x}\sin(y)}
\]

\end{problem}

\begin{problem}(Problem 3c)
Find the real and imaginary parts of the function $w = f(z) = e^{\overline{z}}$
\[
u(x,y) = \answer{e^x\cos(y)} \;\; \mbox{and} \;\; v(x,y) = \answer{-e^x\sin(y)}
\]

\end{problem}



\section{The Principal Argument Function}

Recall the polar form of a complex number:
\[
z = r\cis\theta
\]
where $r = |z|$ and $\theta$ is an angle co-terminal with the vector from $0$ to $z$. Such an angle is called an {\bf argument}
of the complex number. If $\theta$ is an argument of $z$, then any angle of the form $\theta +2k\pi$ where $k$ is an integer, is also an argument of $z$.

We define the set
\[
\arg(z) = \{\theta + 2k\pi | k \in \mathbb{Z}\}
\]
where $\theta$ is any argument of $z$.
To create an argument function, we must select one member of the set $\arg(z)$ for each $z \in \mathbb{C}$.
The principal argument function, $\Arg(z), z \neq 0$, is defined by
\[
\Arg(z) = \theta, \;\mbox{where}\; \theta\in\arg(z)\;\mbox{and}\; -\pi < \theta \leq \pi
\]

\begin{remark}
$\Arg(0)$ is not defined.
\end{remark}

\begin{example}[Example 4] 
Let $w = f(z) = \Arg(z)$. Find $u(x,y)$ and $v(x,y)$.\\ 
First, since $\Arg(z)$ is a real number, we have $u(x,y) = Arg(z)$ and $v(x,y) = 0$.
To find $u(x,y)$ we use the inverse tangent function.
If $z=x+iy$ is in the first or fourth quadrant, then 
\[
\Arg(z) = \tan^{-1}\left(\frac{y}{x}\right).
\]
If $z$ is in the second quadrant, then 
\[
\Arg(z) = \tan^{-1}\left(\frac{y}{x}\right) + \pi.
\]
If $z$ is in the third quadrant, then 
\[
\Arg(z) = \tan^{-1}\left(\frac{y}{x}\right) - \pi.
\]
If $z =x$ is a real number, then $Arg(z) = 0$ if $x>0$ and  $Arg(z) = \pi$ if $x<0$.\\
If $z = iy$ is a purely imaginary number, then $Arg(z) = \pi/2$ if $y>0$ and $Arg(z) = -\pi/2$ if $y<0$.
\end{example}


\begin{problem}(Problem 4a)
Find the following principal arguments:\\
$i) \Arg(\pi) = \answer{0}$\\
$ii) \Arg(-2\pi) = \answer{\pi}$\\
$iii) \Arg(\pi i) = \answer{\pi/2}$\\
$iv) \Arg(-i) = \answer{-\pi/2}$\\
$v) \Arg(1-i) = \answer{-\pi/4}$\\
$vi) \Arg(-1+i\sqrt 3) = \answer{2\pi/3}$
\end{problem}

\begin{problem}(Problem 4b)
Recall that a geometric property of complex multiplication is the addition of angles:
\[
z_1 \cdot z_2 = r_1 \cis\theta_1 \cdot r_2 \cis\theta_2 = r_1r_2\cis\left(\theta_1+\theta_2\right)
\]
Now, let's consider $\Arg(z_1 \cdot z_2)$ and ruminate on the following question.
Does the equation 
\[
\Arg(z_1\cdot z_2) = \Arg(z_1) + \Arg(z_2)
\]
hold for all non-zero complex numbers $z_1$ and $z_2$?
\begin{multipleChoice}
\choice{Yes}
\choice[correct]{No}
\end{multipleChoice}
Which of the following pairs of complex numbers can be used as a counter-example?
\begin{selectAll}
\choice{$1+i$ and $i-1$}
\choice{$1+i$ and $1-i$}
\choice[correct]{$1+i$ and $-1$}
\choice[correct]{$-i$ and $-1-i$}
\end{selectAll}

\end{problem}



\section{Complex Roots}

The Principal Argument function plays a role in creating a Principal $n^{th}$ root function.
We begin with the square root. Just as there are two square roots of a positive real number, there are two square roots of a non-zero complex number.
For the real square root, we distinguish these roots as positive and negative. However, this concept does not apply to non-real 
complex numbers. To find roots, we use the polar form $z = r\cis\theta = |z|e^{i\theta}$.

\begin{example}[Example 5] 
Find the two square roots of $4i$. \\
The modulus of $4i$ is $4$ and the Principal Argument is $\pi/2$, so in polar form
\[
4i = 4\cis(\pi/2) = 4e^{\pi i/2}
\]
Since squaring a complex number squares the modulus and doubles the argument, 
to find a square root, we should take the (positive) square root of the modulus and halve the argument.
Thus
\[
\sqrt{4i} = 2\cis(\pi/4) = 2e^{\pi i/4} = \sqrt 2 + i\sqrt 2
\]
The other square root is the negative of this one, but we will compute it directly using a different argument for $4i$.
The idea of this method generalizes to $n^{th}$ roots.
We can also write $4i$ in polar form as $4\cis(5\pi/2)$.  Thus the other square root is
\[
\sqrt{4i} = 2\cis(5\pi/4) = -\sqrt 2 - i\sqrt 2
\]

\end{example}


\begin{definition}[Prinicpal Square Root]
If $z \neq 0$, the function
\[
f(z) = z^{1/2} =  |z|^{1/2} \cis\left(\frac{\Arg z}{2}\right) = |z|^{1/2} e^{i\left(\frac{\Arg z}{2}\right)}
\]
is called the Principal Square Root function.
\end{definition}


\begin{remark} In the preceding example, since $\Arg 4i = \pi/2$, the Principal Square Root is
\[
(4i)^{\frac12} = 2\cis(\pi/4) = 2e^{\pi i /4} = \sqrt 2 + i\sqrt 2
\] 
\end{remark}


\begin{problem}(Problem 5a)
Find the Prinicipal Square Root of the complex number. Write your answer in the form $a + bi$.\\
i) $(2i)^{\frac12} = \answer{1+i}$\\
ii) $(-144)^{\frac12} = \answer{12i}$\\
iii) $(2 + 2i\sqrt 3)^{\frac12} = \answer{\sqrt 3 + i}$\\
%4 cis 60 ^1/2 = 2 cis 30
iv) $(-18+ 18i\sqrt 3)^{\frac12} = \answer{3 + 3i\sqrt 3}$\\
% 36 cis 120 ^1/2 = 6 cis 60
\end{problem}


The Principal $n^{th}$ Root function is defined by taking the positive $n^{th}$ root of the modulus and 
$(1/n)^{th}$ of the Principal Argument.

\begin{definition}[Principal $n^{th}$ Root function]
If $n$ is a positive integer and $z \neq 0$, the function
\[
f(z) = z^{\frac1n} =  |z|^{\frac1n} \cis\left(\frac{\Arg z}{n}\right) = |z|^{\frac1n} e^{i\frac{\Arg(z)}{n}}
\]
is called the Principal $n^{th}$ Root function.
\end{definition}


\begin{example}[Example 6]
Find the Principal Cube Root of $-8i$ and then find all of the other cube roots. Include them in a sketch.\\
Since $|-8i| = 8$ and $\Arg(-8i) = -\pi/2$, we have
\[
(-8i)^{\frac13} = 8^{\frac13} \cis(-\pi/6) = \sqrt 3 - i
\]
To find the other cube roots, we add multiples of $2\pi$ to the argument before dividing by $3$.
This gives
\begin{align*}
\sqrt[3]{-8i} &= 2 \cis\left[\frac13\left(\frac{-\pi}{2} + 2k\pi\right)\right] \\
              &= 2 \cis\left(-\frac{\pi}{6} + \frac{2k\pi}{3}\right)
\end{align*}
     This gives distinct values if $k = 0, 1$ or $2$. If $k=0$ we get the Principal Cube Root.
     If $k = 1$ we get
\[
\sqrt[3]{-8i}= 2 \cis\left(\frac{\pi}{2}\right)  = 2i 
\]
and if $k = 2$, we get
\[
\sqrt[3]{-8i}= 2 \cis\left(\frac{7\pi}{6}\right) = -\sqrt 3 - i 
 \]    
\begin{image}
\begin{tikzpicture}
\draw [blue, <->](-2.5, 0) -- (2.5,0);
\draw[blue, <->] (0,2.5) -- (0, -2.5) node[ below=10pt, blue]{\large The Cube Roots of $-8i$  };
%\node @(0,-2.5){The three cube roots of $-8i$}
\draw[mark=*,mark size=1pt,mark options={color=blue}] plot coordinates {(0,2)} node[right, blue]{$2i$};
\draw[mark=*,mark size=1pt,mark options={color=red}] plot coordinates {(1.732,-1)} node[below right, red]{$\sqrt 3 - i $};
\draw[mark=*,mark size=1pt,mark options={color=blue}] plot coordinates {(-1.732,-1)} node[below left, blue]{$-\sqrt 3 - i $};
\draw [red, dashed](0, 0) -- (1.732, -1);
\draw [blue, dashed](0, 0) -- (-1.732, -1);
\draw [blue, dashed, thick](0, 0) -- (0,2);
%\node (test) [rectangle, draw] {this node \\ has \\ four \\lines};(Principal Cube Root in red)
\end{tikzpicture}
\end{image}

\end{example}


\begin{problem}(problem 6a)
Find and sketch the 3 cube roots of each of the following:\\
\begin{align*}
i) & \;1 \; \mbox{(these are called roots of unity)} \\
ii)& \;27 \\
iii) & \;-8 \\
iv) & \;i
\end{align*}
\end{problem}

\begin{problem}(problem 6b)
Find and sketch the 4 fourth roots of each of the following:\\
\begin{align*}
i) &\; 1 \; \mbox{(these are called roots of unity)} \\
ii)&\; 16 \\
iii) &\; -81 \\
iv) & \;i
\end{align*}
\end{problem}

\begin{problem}(problem 6c)
Find and sketch the 5 fifth roots of each of the following:\\
\begin{align*}
i) & \;1 \; \mbox{(these are called roots of unity)} \\
ii)& \;-32 \\
iii) & \;i
\end{align*}
\end{problem}

\end{document}

 












Consider the function $w = f(z) = xy +i(x+y)$.  Expressing this function directly in terms of $u(x,y)$ and $v(x,y)$ masks
a direct formula for the function in terms of $z$.  
However, we can always think of $x$ as $\Rep(x)$ and $y$ as $\Imp(z)$ and thereby express $f$ in terms of these.
This is a bit clunky and better is to note that 
\[
x= \frac{z+\overline{z}}{2} \;\; \mbox{and} \;\; y= \frac{z-\overline{z}}{2i}
\]
Using $z$ and $\overline{z}$ we can write the above function as
\[
w=f(z) = \frac{z+\overline{z}}{2} \cdot \frac{z-\overline{z}}{2i} + i\left(\frac{z+\overline{z}}{2} + \frac{z-\overline{z}}{2i}\right)
\]
which simplifies to $w = f(z) = \frac{z^2-\overline{z}^2}{4i} + i\frac{z-\overline{z}}{2}







\begin{center}
\begin{foldable}
\unfoldable{Here is a video of Example 1}
\youtube{###} %vid of example 1
\end{foldable}
\end{center}

\begin{problem} (Problem 1)
  Problem statement
  
    \begin{hint}
      Hint goes here
    \end{hint}
    
    
		
		answer box:
		 $\answer[given]{number}$

\end{problem}







\begin{example}[Example 15]

\end{example}

%more question formats below




%\begin{verbatim}
\begin{question}
What is your favorite color?
\begin{multipleChoice}
\choice[correct]{Rainbow}
\choice{Blue}
\choice{Green}
\choice{Red}
\end{multipleChoice}
\begin{freeResponse}
Hello
\end{freeResponse}
\end{question}
%\end{verbatim}





\begin{question}
  Which one will you choose?
  \begin{multipleChoice}
    \choice[correct]{I'm correct.}
    \choice{I'm wrong.}
    \choice{I'm wrong too.}
  \end{multipleChoice}
\end{question}


\begin{question}
  Which one will you choose?
  \begin{selectAll}
    \choice[correct]{I'm correct.}
    \choice{I'm wrong.}
    \choice[correct]{I'm also correct.}
    \choice{I'm wrong too.}
  \end{selectAll}
\end{question}


\begin{freeResponse}
What is the chain rule used for?
\end{freeResponse}
