\documentclass[handout]{ximera}

\input{../preamble}

\outcome{In this section we compute triple integrals over various regions.}

\title{4.4 Triple Integrals}



\begin{document}

\begin{abstract}
In this section we compute triple integrals over various regions.
\end{abstract}
 
\maketitle

A triple integral is computed in a manner similar to a double integral.  Instead of integrating over a region in the $xy$-plane,
we integrate over regions in $xyz$-space.  Thus, triple integrals will become three iterated integrals (rather than two).
For simplicity, we will assume that our regions include boundaries of the variable $z$ of the form 
\[
g_1(x,y) \leq z \leq g_2(x,y)
\]
and we will integrate either $dz\, dy\, dx$ or $dz\, dx\, dy$.


\begin{example}[Example 1]
Compute $\iiint_R xyz \, dV$ where $R$ is the rectangular region bounded by the 
planes $x = 0, x = 3, y = 1, y = 2, z = -2$ and $z = 3$.\\
The region $R$ is rectangular with constant boundaries for each of the three variables.  Furthermore, the integrand
is a separable function, i.e., $f(x,y,z) = xyz = f_1(x) \cdot f_2(y) \cdot f_3(z)$. Hence, our iterated integrals can 
be written as a product of three ordinary definite integrals. We have:

\begin{align*}
\iiint_R xyz \, dV & = \int_0^3\int_1^2 \int_{-2}^3 xyz \,dz\, dy \, dx\\
                 & =\left(\int_0^3 x \, dx \right) \cdot \left(\int_1^2 y\, dy\right) \cdot \left(\int_{-2}^3z \,dz \right)\\
                 & = \frac92 \cdot \left(\frac{4}{2} -\frac12\right) \cdot \left(\frac92 - \frac42\right)\\
                 &=  \frac{135}{8}
\end{align*}      

\end{example}



\begin{example}[Example 2]
Compute $\iiint_R x \, dV$ where $R$ is the tetrahedron in the first octant with 
vertices $(0,0,0), P(1, 0, 0), Q(0, 2, 0)$ and $R(0,0,3)$.


The plane determined by the points $P, Q$ and $R$ has normal vector $\vec{PQ} \cross \vec{PR} = \vector{6, 3, 2}$
and has equation 
\[
6x + 3y + 2z = 6
\]
Hence, the boundaries for $z$ come from
\[
0 \leq z \leq \frac{6 - 6x - 3y}{2}
\]

The base of the tetrahedron is the triangle in the $xy$-plane with vertices $(0,0), (1,0)$ and $(2,0)$. 
The line connecting the points $(1,0)$ and $(0,2)$ has equation $y = -2x + 2$.
Thus, the base of the tetrahedron can be described using the inequalities
\[
\quad 0 \leq y \leq -2x+2, \quad 0 \leq x \leq 1
\]
The triple integral over the tetrahedron can now be written as iterated integrals and computed:
\begin{align*}
\iiint_R x \, dV & = \int_0^1 \int_0^{-2x+2} \int_0^{3 - 3x - 3y/2} x \, dz\, dy \, dx\\
                 & = \int_0^1 \int_0^{-2x+2}  xz\bigg|_0^{3 - 3x - 3y/2} \, dy \, dx\\
                 & = \int_0^1 \int_0^{-2x+2}  (3x - 3x^2 - 3xy/2) \, dy \, dx\\
                 &= \int_0^1   (3xy - 3x^2y - 3xy^2/4)\bigg|_0^{-2x+2} \, dx\\
                 &= \int_0^1 3x(-2x+2) -3x^2(-2x+2) -3x(-2x+2)^2/4 \, dx\\
                 &= \int_0^1   (3x^3 - 6x^2  +  3x)   \, dx\\
                 &= \frac34 - 2 + \frac32 = \frac14
\end{align*}  

\end{example}



\begin{example}[Example 3]
Compute $\iiint_R  e^{z/y} \, dV$ where $R$ is the region in the first octant below the 
surface $z = xy$ and above the triangle in the $xy$-plane with vertices $(0,0), (1,0)$ and $(1,1)$.\\


The boundaries for $z$ are $0 \leq z \leq xy$. The boundaries for $x$ and $y$ come from the triangle bounded by the lines
$y = 0, x = 1$ and $y = x$.  We choose to integrate with respect to $x$ and then with respect to $y$ because the
integrand is (slightly) easier to deal with.  This detail is not evident until after we integrate with respect to $z$.
We have:

\begin{align*}
\iiint_R e^{z/y} \, dV & = \int_0^1 \int_y^1 \int_0^{xy} e^{z/y} \, dz \, dx \, dy\\
                 & = \int_0^1 \int_y^1 ye^{z/y}\bigg|_{0}^{xy} \,dx \, dy\\
                 & = \int_0^1 \int_y^1 \left(ye^x - y\right) \,dx \, dy\\
                 &= \int_0^1 \left(ye^x - xy\right) \bigg|_y^1 \, dy\\
                 &= \int_0^1 \left(ye - y - ye^y + y^2\right) \, dy\\
                 &= \frac{y^2 e}{2} - \frac{y^2}{2} - (y-1)e^y + \frac{y^3}{3} \bigg|_0^1\\
                 &= \frac{e}{2} - \frac12 + \frac13 - (-1)\\
                 &= \frac{e-1}{2} - \frac23
\end{align*} 

\end{example}



\end{document}
