\documentclass{ximera}

\input{../preamble}

\outcome{Find the Maclaurin Series representation of a function}

\title{Maclaurin Series}

\begin{document}

\begin{abstract}
We find the Maclaurin Series representation of a function.
\end{abstract}

\maketitle

\section{Maclaurin Series}


In this section we will examine a special type of Taylor Series, called a Maclaurin Series and we will develop the Maclaurin Series for 
some important functions.

\begin{definition}[Maclaurin series] 
A \textbf{Maclaurin Series} is a Taylor Series with center 0, i.e., a power series representation 
for a function, $f(x)$, of the form
\[
f(x) = \sum_{n=0}^\infty c_n x^n.
\]
\end{definition}
 
We have already seen the series representation for the function $f(x) = \frac{1}{1-x}$,
\[
\frac{1}{1-x} = \sum_{n=0}^\infty  x^n.
\]
In the current context, we refer to this series representation as the Maclaurin Series for the function $\frac{1}{1-x}$.
Furthermore, we have seen the formula for the coefficients of a Taylor Series representation centered at $a$,
\[
c_n = \frac{f^{(n)}(a)}{n!},
\]
and we can use this formula to find a Maclaurin Series representation by letting $a = 0$.
We will do this for the functions $e^x, \sin(x),$ and $\cos(x)$.

\begin{example}[example 1 - the Maclaurin Series for $e^x$] 
Find the Maclaurin Series representation for the function $f(x) = e^x$.\\
To find the Maclaurin Series representation, we must determine the coefficients, $c_n$. These are particularly easy to find for the exponential function
since $f^{(n)}(x) = e^x$ for all whole numbers, $n$. Thus the coefficients are,
\[
c_n = \frac{f^{(n)}(0)}{n!} = \frac{e^0}{n!} = \frac{1}{n!},
\]
and the Maclaurin Series representation is 
\[
e^x = \sum_{n=0}^\infty c_n x^n = \sum_{n=0}^\infty \frac{x^n}{n!} = 1 + x + \frac{x^2}{2!} + \frac{x^3}{3!} + \cdots.
\]
We can use the ratio test to verify that this representation converges for all $x$ and 
hence the representation is valid on the interval $(-\infty, \infty)$.
\end{example}

We now turn to two examples of finding Maclauirin Series by making modifications to the previous example.

\begin{example}[example 2] 
Find the Maclaurin Series representation of the function $f(x) = e^{-x}$.\\
Replacing `$x$' in the Maclaurin Series representation for $e^x$ with $-x$ gives
\[
e^{-x} = \sum_{n=0}^\infty \frac{(-x)^n}{n!} = \sum_{n=0}^\infty (-1)^n \frac{x^n}{n!} = 1 - x + \frac{x^2}{2!} - \dots.
\]
This representation is valid on the interval $(-\infty, \infty)$.
\end{example}


\begin{example}[example 3] 
Find the Maclaurin Series representation of the function $f(x) = x^2 e^{2x}$.\\
Replacing `$x$' with $2x$ in the Maclaurin Series representation for $e^x$ gives
\[
x^2 e^{2x} = x^2\sum_{n=0}^\infty \frac{(2x)^n}{n!} = \sum_{n=0}^\infty x^2 \cdot \frac{2^n x^n}{n!} = \sum_{n=0}^\infty \frac{2^n x^{n+2}}{n!}.
\]
This representation is valid on the interval $(-\infty, \infty)$.
\end{example}


\begin{example}[example 4 - the Maclaurin Series for $\sin(x)$] 
Find the Maclaurin Series representation for $f(x) = \sin(x)$.\\
To find the Maclaurin series representation, we must determine the coefficients, $c_n$. 
We will take advantage of a cyclical pattern in the derivatives of $\sin(x)$.
The first four derivatives are:
\[
f'(x) = \cos(x), f''(x) = -\sin(x), f'''(x) - -\cos(x) \text{ and } f^{(4)}(x) = \sin(x).
\]
Since we arrived back at $\sin(x)$, the next four derivatives will be exactly the same.
The coefficients require us to plug in the center, $x = 0$:
\[
f(0) = \sin(0) = 0, f'(0) = \cos(0) = 1,
\]
\[
 f''(0) = -\sin(0) = 0, \text{ and } f'''(0) = -\cos(0) = -1.
\]
Hence the numerators of the coefficients will cycle through the values $0, 1, 0,$ and $-1$.
Since the numerators of the even coefficients are 0, these coefficients are themselves 0 (because $0/n! = 0$) for any whole number, $n$.
The numerators of the odd coefficients will alternate between $1$ and $-1$ and hence 
\[
c_1 = \frac{1}{1!}, c_3 = \frac{-1}{3!}, c_5 = \frac{1}{5!}, c_7 = \frac{-1}{7!}, \text{etc}.
\]

We can now write the Maclaurin Series representation as
\[
\sin(x) =  x - \frac{x^3}{3!} + \frac{x^5}{5!} - \frac{x^7}{7!} + \cdots.
\]
This can be writtien in summation notation as
\[
\sin(x) = \sum_{n=0}^\infty \frac{(-1)^{n+1}}{(2n+1)!}x^{2n+1}.
\]
The ratio test can be used to verify that this representation is valid on the interval $(-\infty, \infty)$.
\begin{remark}
Recall that $\sin(x)$ is an \textit{odd} function, 
i.e., $\sin(-x) = -\sin(x)$, and now notice that the Maclaurin Series representation for $\sin(x)$ consists of only the 
odd powers of $x$. 
\end{remark}
\end{example}


We now consider an example of a Maclaurin Series obtained by making modifications to the previous example.


\begin{example}[example 5]
Find the Maclaurin Series representation for the function $f(x) = x\sin(3x)$.\\
We will use the Maclaurin Series for $\sin(x)$ with $3x$ replacing $x$ and then multiply the result by $x$.
This gives
\[
x\sin(3x) = x \sum_{n=0}^\infty \frac{(-1)^{n+1}}{(2n+1)!}(3x)^{2n+1} = \sum_{n=0}^\infty (-1)^{n+1} \frac{3^{2n+1}}{(2n+1)!}x^{2n+2}.
\]
This representation is valid in the interval $(-\infty, \infty)$.
\end{example}


\begin{example}[example 6 - the Maclaurin Series for $\cos(x)$] 
Find the Maclaurin Series representation for the function $f(x) = \cos(x)$.\\
Most of the work has already been done for us in the $\sin(x)$ example.  The numerators of the coefficients will cycle through the same
four values as $\sin(x)$, but instead of starting with 0, they start with $\cos(0) = 1$. 
Specfically, the numerators of the coefficients cycle through the values $1, 0, -1$ and $0$.  When the numerator is 0, the coefficient is 0,
so the Maclaurin Series representation for $\cos(x)$ will only consist of even powers of $x$:
\[
\cos(x) = 1 + 0x - \frac{x^2}{2!} + 0x^3 + \cdots = 1 - \frac{x^2}{2!} + \frac{x^4}{4!} - \frac{x^6}{6!} + \cdots.
\]
This can be writtien in summation notation as
\[
\cos(x) = \sum_{n=0}^\infty \frac{(-1)^n}{(2n)!}x^{2n}.
\]
The ratio test can be used to verify that this representation is valid on the interval $(-\infty, \infty)$.
\begin{remark}
Recall that $\cos(x)$ is an \textit{even} function, 
i.e., $\cos(-x) = \cos(x)$, and as mentioned the Maclaurin Series representation for $\cos(x)$ consists of only the 
even powers of $x$.
\end{remark} 
\end{example}


We now consider an example of a Maclaurin Series obtained by making modifications to the previous example.

\begin{example}[example 7]
Find the Maclaurin Series representation for the function 
\[
f(x) = x^2\cos(2x).
\]
We will use the Maclaurin Series for $\cos(x)$ with $2x$ replacing $x$ and then multiply the result by $x^2$.
This gives
\[
x^2\cos(2x) = x^2\sum_{n=0}^\infty \frac{(-1)^n}{(2n)!}(2x)^{2n} = \sum_{n=0}^\infty (-1)^n\frac{2^{2n}}{(2n)!}x^{2n+2}.
\]
This representation is valid in the interval $(-\infty, \infty)$.
\end{example}


\section{Video Lessons}


\begin{center}
\begin{foldable}
\unfoldable{Here is a detailed, lecture style video on Maclaurin Series:}
\youtube{aa9rcfNJrfk}
\end{foldable}
\end{center}





\end{document}


\begin{example} %example #15
Find $h'(x)$ if $h(x) = x^{\sin(x)}$.\\
We will use the fact that the exponential and logarithm functions are inverses,
\[e^{\ln(x)} = x,\]
and the exponent property of logarithms, 
\[\ln(x^n) = n\ln(x),\]
to rewrite $h(x)$.  We have 
\[h(x) = x^{\sin(x)} = e^{\ln(x^{\sin(x)})} = e^{\sin(x)\ln(x)}\]
and we can now compute $h'(x)$ using a combination of the chain rule and product rule.
We can write $h(x)$ as a composition, $f(g(x))$ with 
\[g(x) = \sin(x)\ln(x) \quad \text{and} \quad f(x) = e^x.\]
Then to find $g'(x)$ we us the product rule and we get $g'(x) = \frac{\sin(x)}{x} + \cos(x)\ln(x)$.
Next $f'(x) = e^x$ and 
hence $f'(g(x)) = e^{g(x)} = e^{\sin(x)\ln(x)} = x^{\sin(x)}$.
We can then conclude $h'(x) = f'(g(x))g'(x) = x^{\sin(x)} \left[ \frac{\sin(x)}{x} + \cos(x)\ln(x)\right]$.
\end{example}

%more question formats below













%\begin{verbatim}
\begin{question}
What is your favorite color?
\begin{multipleChoice}
\choice[correct]{Rainbow}
\choice{Blue}
\choice{Green}
\choice{Red}
\end{multipleChoice}
\begin{freeResponse}
Hello
\end{freeResponse}
\end{question}
%\end{verbatim}





\begin{question}
  Which one will you choose?
  \begin{multipleChoice}
    \choice[correct]{I'm correct.}
    \choice{I'm wrong.}
    \choice{I'm wrong too.}
  \end{multipleChoice}
\end{question}


\begin{question}
  Which one will you choose?
  \begin{selectAll}
    \choice[correct]{I'm correct.}
    \choice{I'm wrong.}
    \choice[correct]{I'm also correct.}
    \choice{I'm wrong too.}
  \end{selectAll}
\end{question}


\begin{freeResponse}
What is the chain rule used for?
\end{freeResponse}
