\documentclass[handout]{ximera}

\input{../preamble}

\pgfplotsset{compat=1.13}

\outcome{Learn the basics of generating functions}

\title{3.2 Generating Functions}

\begin{document}

\begin{abstract}
We will define and interpret generating functions.
\end{abstract}

\maketitle

\section{Generating Functions}

\begin{definition}
A \textbf{generating function} is a function of the form
\[
g(x) = \sum_{k=0}^\infty c_kx^k = c_0 + c_1 x + c_2 x^2 + \cdots
\]
whose coefficients $c_0, c_1, c_2, ...$ encode information related to a counting problem.
\end{definition}


\begin{example}[example 1]
The function
\[
g_n(x) = \sum_{k=0}^n \binom{n}{k}x^k  = \binom{n}{0}+ \binom{n}{1}x + \binom{n}{2}x^2 + \cdots + 
\binom{n}{k}x^k + \cdots + \binom{n}{n} x^n
\]
The coefficients $c_{n+1}, c_{n+2}, ...$ are all zero.
The non-zero coefficients are binomial coefficients- the coefficient of $x^k$ is the number of subsets of an $n$-element set.
\end{example}

\begin{problem}
Consider the generating function
\[
g_n(x) = \sum_{k=0}^n \frac{n!}{(n-k)!}x^k  = \frac{n!}{n!}+ \frac{n!}{(n-1)!}x + \frac{n!}{(n-2)!}x^2 + \cdots + 
\frac{n!}{(n-k)!}x^k + \cdots + \frac{n!}{(n-n)!} x^n
\]
Give an example of a counting problem whose solution is given by the coefficients of $g_n(x)$.
\end{problem}

\begin{example}[example 2]
Flip a coin and consider the function $g(x) = 1+x$.  The coefficient of $x^k, k = 0,1$ represents the number ways of 
obtaining $k$ ``heads". Now flip the coin twice and consider the function $g_2(x) = (1+x)^2 = 1 + 2x + x^2$.
Note that there is one way to get zero heads: TT. There are two ways to get one heads, HT and TH, 
and there is one way to get two heads, HH. Thus the function $g_2(x)$ encodes then number of ways to 
get $k$ heads in two flips of the coin, 
where $k = 0, 1, 2$.
\end{example}

\begin{problem}(problem 2a)
Based on the example above guess the generating function corresponding to counting the number of heads in 3 coin flips.
Express your guess both as a power of $1+x$ as a proper generating function with numerical coefficients.
Verify that your guess is correct. 
\end{problem}

\begin{problem}(problem 2b)
Generalize your answer to the last problem by creating the generating function corresponding to the 
number of heads in $n$ coin flips.
Express your answer as a power of $1+x$ and verify your answer using the Binomial Theorem.
\end{problem}
 

\begin{example}[example 3]
Consider a six-sided die whose faces are numbered 1 through 6. If it is rolled once, the function
\[
g_1(x) = x + x^2 + x^3 + x^4 + x^5 + x^6
\]
encodes the number of ways of obtaining the outcome corresponding to the degree of a particular term.
Now roll the die twice and consider the sum. It can be $2, 3, 4, ..., 11, 12$. Squaring $g_1(x)$ gives
\[
g_2(x) = x^2 + 2x^3 + 3x^4 + 4x^5 + 5x^6 + 6x^7 + 5x^8 + 4x^9+ 3x^{10} + 2x^{11} + x^{12} \;\; \text{verify}
\]
The coefficient of $x^k$ in $g_2(x)$ gives the number of ways of rolling a sum of $k$ (verify).
\end{example}

\begin{problem}(problem 3)
Consider a six-sided die whose faces are numbered 1 through 6. Roll the die 3 times and consider the sum.
What are the minimum and maximum possible values of the sum? Create a generating function $g_3(x)$ whose coefficients 
encode the the number of ways of rolling a sum of $k$.  Express your answer as a power of $g_1(x)$ from the example above.
Use a computer algebra system to find the coefficient of $x^7$. Verify that this coefficient is 
indeed the number of ways of obtaining a sum of $7$ by enumerating the possibilities.
\end{problem}


\begin{example}[example 4]
Consider the number of non-negative integer solutions of the equation
\[
x_1 + x_2 + \cdots + x_k = n
\]
Find the generating function $g_k(x)$ whose coefficients give the number of such solutions.\\
Each variable $x_i$ can take on any value (up to the maximum possible $n$, which is unspecified) and so it will contribute 
a factor of 
\[
1 + x + x^2 +x^3 + \cdots = \frac{1}{1-x}
\]
Hence
\[
g_k(x) = (1+x+x^2 + \cdots)^k = \frac{1}{(1-x)^k}
\]
\end{example}

\end{document}







